### 2018-02-13 paper.Rmd
First commit

---
output:
  pdf_document: default
  word_document: default
  html_document: default
---

```{r setup, include=FALSE, echo=FALSE}
library(ggplot2)
library(plyr)
library(sp)
library(rgdal)
library(ggalt)
library(dplyr)

long_norway <- c(0, 32)
lat_norway <- c(57, 72)

```

```{r exploration, include=FALSE, echo=FALSE}
source("exploration.R")
```

```{r author_info, include=FALSE, echo=FALSE}
first_name <- "Wouter"
last_name <- "Koch"
institution <- "Norwegian Biodiversity Information Centre"
address <- "Trondheim, Norway"
```

```{r read_ao_data, include=FALSE, echo=FALSE}
ao_data <- read_gbif(path="data/GBIF_Artsobs_all_2016.csv")
```

# Norwegian citizen science data exploration

By `r last_name`, `r first_name`

> `r institution`, `r address`

---------------------------------------------

```{r filter_data, include=FALSE, echo=FALSE}
genera <- c("Larus")
ao_data_subset <- filter_by_genus(data=ao_data, genera=genera)

long_dataset <- c(min(ao_data_subset$decimallongitude) %/% 1 , max(ao_data_subset$decimallongitude) %/% 1 + 1)
lat_dataset <- c(min(ao_data_subset$decimallatitude) %/% 1 , max(ao_data_subset$decimallatitude) %/% 1 + 1)
# long_dataset <- long_norway
# lat_dataset <- lat_norway

```

A test data exploration of some [Norwegian Species Observation Service](https://www.artsobservasjoner.no) data, obtained through [GBIF](https://www.gbif.org) (N = `r nrow(ao_data)`).

Filtering on `r summarize_list(genera)` leaves `r nrow(ao_data_subset)` observations, plotted on the map below.

```{r map, echo=FALSE, message=FALSE}
world_map <- map_data("world", "Norway")

ggplot(ao_data_subset) + annotation_map(world_map, fill="gray50", colour="gray50") +
    geom_point(aes(x=decimallongitude, y=decimallatitude, color=species), alpha=0.3, size = .3) +
    theme(panel.background = element_rect(fill="aliceblue"), legend.position = "right") +
    coord_map(
      projection = "mercator",
      xlim=long_dataset,  # limits on longitude
      ylim=lat_dataset) + # limits on latitude
    xlab("Longitude") +
    ylab("Latitude")

```

Map for `r summarize_list(genera)` in Norway

```{r rasterize, message=FALSE, echo=FALSE}

  gridsizeinkm <- 50

  ao_data_subset <- add_rounded_coordinates(dataset=ao_data_subset, km=gridsizeinkm)
```

```{r rastermap, message=FALSE, echo=FALSE}
world_map <- map_data("world", "Norway")

ggplot(ao_data_subset) + 
    annotation_map(world_map, fill="gray50", colour="gray50") +
    geom_point(aes(x=long_rounded, y=lat_rounded, color=species), alpha=0.3, size = gridsizeinkm * .04) +
    theme(panel.background = element_rect(fill="aliceblue"), legend.position = "right") +
    coord_map(projection = "gnomonic", orientation=c(60, 16, -5),
              xlim=long_dataset,  # limits on longitude
              ylim=lat_dataset) + # limits on latitude
   xlab("Longitude") + ylab("Latitude")

```

Map for `r summarize_list(genera)` in Norway, rasterized to `r gridsizeinkm` x `r gridsizeinkm` km  cells.

```{r densitymap, message=FALSE, echo=FALSE}

counts <- ddply(ao_data_subset, .(ao_data_subset$long_rounded, ao_data_subset$lat_rounded), nrow)
names(counts) <- c("long_rounded", "lat_rounded", "freq")

world_map <- map_data("world", "Norway")

ggplot(ao_data_subset, aes(decimallongitude, decimallatitude)) + 
    annotation_map(world_map, fill="white", color=NA) +
    geom_bin2d(aes(fill = log10(..count..)), binwidth = c(.018 * gridsizeinkm ,  .01  * gridsizeinkm)) +
    scale_fill_gradient(low = "gray90", high = "red") +
    annotation_map(world_map, fill=NA, color="black") +
    theme(panel.background = element_rect(fill="aliceblue"), legend.position = "right") +
    coord_map(projection = "gnomonic", orientation=c(60, 16, -5),
      xlim=long_dataset,  # limits on longitude
      ylim=lat_dataset) + # limits on latitude
   xlab("Longitude") + ylab("Latitude")

```

Map for `r summarize_list(genera)` in Norway, raterized to `r .018 * gridsizeinkm` x `r .01  * gridsizeinkm` degree cells.

## Human population density

```{r human_pop_data, messages=FALSE, echo=FALSE}
geostat_data <- read_geostat(path="data/GeoStat_NO.csv")
geostat_data <- geostat_add_utm(geostat_data)

ggplot(geostat_data) + 
    annotation_map(world_map, fill="gray") +
    geom_point(aes(x=long, y=lat, color=log(X.TOT_P.)), size = .0004, shape=15) +
    scale_color_distiller(palette = "Spectral") +
    annotation_map(world_map, fill=NA, colour="gray30") +
    theme(panel.background = element_rect(fill="aliceblue"), legend.position = "right") +
    coord_map(projection = "gnomonic", orientation=c(60, 16, -5),
              xlim=long_dataset,  # limits on longitude
              ylim=lat_dataset) + # limits on latitude
   xlab("Longitude") + ylab("Latitude")

```

Log human population per square km in norway

```{r human_pop_rasterize, messages=FALSE, echo=FALSE}
geostat_data <- utm_add_rasterized(geostat_data, km=gridsizeinkm)
```

```{r human_pop_plot_rasterized, messages=FALSE, echo=FALSE}
ggplot(geostat_data) + 
    annotation_map(world_map, fill="gray") +
    geom_point(aes(x=long_rounded, y=lat_rounded, color=log(pop)), size = gridsizeinkm * .04) +
     scale_color_distiller(palette = "Spectral") +
    annotation_map(world_map, fill=NA, colour="gray30") +
    theme(panel.background = element_rect(fill="aliceblue"), legend.position = "right") +
    coord_map(projection = "gnomonic", orientation=c(60, 16, -5),
              xlim=long_dataset,  # limits on longitude
              ylim=lat_dataset) + # limits on latitude
   xlab("Longitude") + ylab("Latitude")

```

The log human population per `r gridsizeinkm` x `r gridsizeinkm` km cs cell in Norway.

```{r per_capita_scatterplot, messages=FALSE, echo=FALSE}
  num_vs_pop <- join_by_utm_rounded(ao_data_subset, geostat_data)

  ggplot(num_vs_pop, aes(x=pop, y=num)) + 
    geom_point()+
    geom_smooth(method=lm)
```

The number of `r summarize_list(genera)` in Norway per `r gridsizeinkm` x `r gridsizeinkm` km cs cells, versus the human population in those cells. R² of the linaer regression is `r summary(lm(num ~ pop, data=num_vs_pop))$r.squared`. R² of the log linaer regression is `r summary(lm(log10(num) ~ log10(pop), data=num_vs_pop))$r.squared`.

```{r per_capita_plot_rasterized, messages=FALSE, echo=FALSE}

num_vs_pop$percapita <- num_vs_pop$num / num_vs_pop$pop

ggplot(num_vs_pop) + 
    annotation_map(world_map, fill="gray") +
    geom_point(aes(x=long_rounded, y=lat_rounded, color=log(percapita)), size = gridsizeinkm * .04) +
     scale_color_distiller(palette = "Spectral") +
    annotation_map(world_map, fill=NA, colour="gray30") +
    theme(panel.background = element_rect(fill="aliceblue"), legend.position = "right") +
    coord_map(projection = "gnomonic", orientation=c(60, 16, -5),
              xlim=long_dataset,  # limits on longitude
              ylim=lat_dataset) + # limits on latitude
   xlab("Longitude") + ylab("Latitude")

```

The log number of `r summarize_list(genera)` observations in Norway per capita within `r gridsizeinkm` x `r gridsizeinkm` km cells.

### 2018-02-13 exploration.R

@@ -1,0 +1,97 @@
+read_gbif <- function(path) {
+  result <- read.table(path, sep="\t", header=TRUE, quote = "", fill=TRUE)
+  result
+}
+
+read_geostat <- function(path) {
+  result <- read.table(path, sep=",", header=TRUE, quote = "", fill=TRUE)
+  result
+}
+
+geostat_add_utm <- function(dataset) {
+  dataset$utm_y <- strtoi(substr(dataset$X.GRD_ID, 6, 9)) * 1000
+  dataset$utm_x <- strtoi(substr(dataset$X.GRD_ID, 11, 14)) * 1000
+  
+  xy <- data.frame(ID = 1:length(dataset[, 1]), X = dataset$utm_x, Y = dataset$utm_y)
+  coordinates(xy) <- c("X", "Y")
+  proj4string(xy) <- CRS("+proj=laea +lat_0=52 +lon_0=10 +x_0=4321000 +y_0=3210000 +ellps=GRS80 +units=m +no_defs")
+  res <- spTransform(xy, CRS("+proj=longlat +datum=WGS84"))
+  res <- as.data.frame(res)
+  
+  dataset$long <- res$X
+  dataset$lat <- res$Y
+
+  return(dataset)
+}
+
+filter_by_genus <- function(dataset, genera) {
+  result <- subset(dataset, genus %in% genera & taxonrank == "SPECIES")
+  result <- droplevels(result)
+  result
+}
+
+summarize_list <- function(list) {
+  if (length(list) > 1) {
+    list[length(list)-1] <- paste(list[length(list)-1], list[length(list)], sep=" and ")
+    list <- list[-length(list)]
+  }
+  paste(list, collapse = ", ")
+}
+
+add_rounded_coordinates <- function(dataset, km=1) {
+  xy <- data.frame(ID = 1:length(dataset[, 1]), X = dataset$decimallongitude, Y = dataset$decimallatitude)
+  coordinates(xy) <- c("X", "Y")
+  proj4string(xy) <- CRS("+proj=longlat +datum=WGS84")
+  res <- spTransform(xy, CRS("+proj=laea +lat_0=52 +lon_0=10 +x_0=4321000 +y_0=3210000 +ellps=GRS80 +units=m +no_defs"))
+  res <- as.data.frame(res)
+  
+  dataset$utm_x <- res$X + 500 * km
[…]

### 2018-02-13 paper.Rmd
Use sample data

@@
 ```{r read_ao_data, include=FALSE, echo=FALSE}
-ao_data <- read_gbif(path="data/GBIF_Artsobs_all_2016.csv")
+ao_data <- read_gbif(path="data/Laridae-NO-2016.csv")
 ```
@@
     theme(panel.background = element_rect(fill="aliceblue"), legend.position = "right") +
-    coord_map(
-      projection = "mercator",
+    coord_map(projection = "gnomonic", orientation=c(60, 16, -5),
               xlim=long_dataset,  # limits on longitude
@@
     annotation_map(world_map, fill="gray50", colour="gray50") +
-    geom_point(aes(x=long_rounded, y=lat_rounded, color=species), alpha=0.3, size = gridsizeinkm * .04) +
+    geom_point(aes(x=long_rounded, y=lat_rounded, color=species), alpha=0.3, size = gridsizeinkm * .02) +
     theme(panel.background = element_rect(fill="aliceblue"), legend.position = "right") +
@@
     annotation_map(world_map, fill="gray") +
-    geom_point(aes(x=long_rounded, y=lat_rounded, color=log(pop)), size = gridsizeinkm * .04) +
+    geom_point(aes(x=long_rounded, y=lat_rounded, color=log(pop)), size = gridsizeinkm * .02) +
      scale_color_distiller(palette = "Spectral") +
@@
     annotation_map(world_map, fill="gray") +
-    geom_point(aes(x=long_rounded, y=lat_rounded, color=log(percapita)), size = gridsizeinkm * .04) +
+    geom_point(aes(x=long_rounded, y=lat_rounded, color=log(percapita)), size = gridsizeinkm * .02) +
      scale_color_distiller(palette = "Spectral") +

### 2018-02-13 exploration.R

@@ -1,5 +1,5 @@
 read_gbif <- function(path) {
-  result <- read.table(path, sep="\t", header=TRUE, quote = "", fill=TRUE)
+  result <- read.table(path, sep="\t", header=TRUE, fill=TRUE)
   result
 }
 

### 2018-02-15 paper.Rmd
Generalize functions, uniform density maps

--- 2018-02-13 paper.Rmd
@@
 ---
+title: "Norwegian citizen science data exploration"
+author: "Wouter Koch, NBIC"
+
 output:
@@
 
-long_norway <- c(0, 32)
-lat_norway <- c(57, 72)
+source("exploration.R")
 
+obs_filter <- c("genus", "Larus")
 ```
 
-```{r exploration, include=FALSE, echo=FALSE}
-source("exploration.R")
+******
+
+```{r read_data, include=FALSE, echo=FALSE}
+ao_data <- read_gbif(path="data/Laridae-NO-2016.csv")
+geostat_data <- read_geostat(path="data/GeoStat_NO.csv")
 ```
 
-```{r author_info, include=FALSE, echo=FALSE}
-first_name <- "Wouter"
-last_name <- "Koch"
-institution <- "Norwegian Biodiversity Information Centre"
-address <- "Trondheim, Norway"
-```
+```{r filter_data, include=FALSE, echo=FALSE}
+ao_data_subset <- filter_observations(ao_data, obs_filter)  # filter observations
 
-```{r read_ao_data, include=FALSE, echo=FALSE}
-ao_data <- read_gbif(path="data/GBIF_Artsobs_all_2016.csv")
-```
-
-# Norwegian citizen science data exploration
-
-By `r last_name`, `r first_name`
-
-> `r institution`, `r address`
-
----------------------------------------------
-
-```{r filter_data, include=FALSE, echo=FALSE}
-genera <- c("Larus")
-ao_data_subset <- filter_by_genus(data=ao_data, genera=genera)
+ao_data_subset <- rasterize(ao_data_subset)   # round all observation positions to km
+geostat_data <- rasterize(geostat_data)       # round all population data positions to km
 
@@
 lat_dataset <- c(min(ao_data_subset$decimallatitude) %/% 1 , max(ao_data_subset$decimallatitude) %/% 1 + 1)
-# long_dataset <- long_norway
-# lat_dataset <- lat_norway
-
 ```
@@
 
-Filtering on `r summarize_list(genera)` leaves `r nrow(ao_data_subset)` observations, plotted on the map below.
+Filtering on `r summarize_list(obs_filter[-1])` leaves `r nrow(ao_data_subset)` observations, plotted on the map below.
 
@@
 ggplot(ao_data_subset) + annotation_map(world_map, fill="gray50", colour="gray50") +
-    geom_point(aes(x=decimallongitude, y=decimallatitude, color=species), alpha=0.3, size = .3) +
+    geom_point(aes(x=decimallongitude, y=decimallatitude, color=species), alpha=0.5, size=cellsize/gridsizeinkm, shape=15) +
     theme(panel.background = element_rect(fill="aliceblue"), legend.position = "right") +
-    coord_map(
-      projection = "mercator",
+    coord_map(projection = "gnomonic", orientation=c(60, 16, -5),
               xlim=long_dataset,  # limits on longitude
@@
 
-Map for `r summarize_list(genera)` in Norway
+Map for `r summarize_list(obs_filter[-1])` in Norway
 
 ```{r rasterize, message=FALSE, echo=FALSE}
-
   gridsizeinkm <- 50
-
-  ao_data_subset <- add_rounded_coordinates(dataset=ao_data_subset, km=gridsizeinkm)
+  cellsize <- (lat_dataset[2] - lat_dataset[1]) * gridsizeinkm * .0016
+  ao_data_subset <- rasterize(ao_data_subset, gridsizeinkm)
 ```
@@
     annotation_map(world_map, fill="gray50", colour="gray50") +
-    geom_point(aes(x=long_rounded, y=lat_rounded, color=species), alpha=0.3, size = gridsizeinkm * .04) +
+    geom_point(aes(x=long_rounded, y=lat_rounded, color=species), alpha=0.3, size = .5 * cellsize) +
     theme(panel.background = element_rect(fill="aliceblue"), legend.position = "right") +
@@
 
-Map for `r summarize_list(genera)` in Norway, rasterized to `r gridsizeinkm` x `r gridsizeinkm` km  cells.
+Map for `r summarize_list(obs_filter[-1])` in Norway, rasterized to `r gridsizeinkm` x `r gridsizeinkm` km  cells.
 
-```{r densitymap, message=FALSE, echo=FALSE}
+```{r densitymap_2, message=FALSE, echo=FALSE}
 
@@
 
-world_map <- map_data("world", "Norway")
-
-ggplot(ao_data_subset, aes(decimallongitude, decimallatitude)) + 
-    annotation_map(world_map, fill="white", color=NA) +
-    geom_bin2d(aes(fill = log10(..count..)), binwidth = c(.018 * gridsizeinkm ,  .01  * gridsizeinkm)) +
-    scale_fill_gradient(low = "gray90", high = "red") +
-    annotation_map(world_map, fill=NA, color="black") +
+ggplot(counts) + 
+    annotation_map(world_map, fill="gray50", colour="gray50") +
+    geom_point(aes(x=long_rounded, y=lat_rounded, color=log(freq)), shape=15, size=cellsize) +
+    scale_color_distiller(palette = "Spectral") +
+    annotation_map(world_map, fill=NA, colour="gray30") +
     theme(panel.background = element_rect(fill="aliceblue"), legend.position = "right") +
@@
    xlab("Longitude") + ylab("Latitude")
-
 ```
 
-Map for `r summarize_list(genera)` in Norway, raterized to `r .018 * gridsizeinkm` x `r .01  * gridsizeinkm` degree cells.
+Map for `r summarize_list(obs_filter[-1])` in Norway, raterized to `r .018 * gridsizeinkm` x `r .01  * gridsizeinkm` degree cells.
 
@@
 ```{r human_pop_data, messages=FALSE, echo=FALSE}
-geostat_data <- read_geostat(path="data/GeoStat_NO.csv")
-geostat_data <- geostat_add_utm(geostat_data)
-
 ggplot(geostat_data) + 
     annotation_map(world_map, fill="gray") +
-    geom_point(aes(x=long, y=lat, color=log(X.TOT_P.)), size = .0004, shape=15) +
+    geom_point(aes(x=long_rounded, y=lat_rounded, color=log(TOT_P)), size=cellsize/gridsizeinkm, shape=15) +
     scale_color_distiller(palette = "Spectral") +
@@
 ```{r human_pop_rasterize, messages=FALSE, echo=FALSE}
-geostat_data <- utm_add_rasterized(geostat_data, km=gridsizeinkm)
+geostat_data_rastered <- rasterize(geostat_data, gridsizeinkm)
 ```
@@
 ```{r human_pop_plot_rasterized, messages=FALSE, echo=FALSE}
-ggplot(geostat_data) + 
-    annotation_map(world_map, fill="gray") +
-    geom_point(aes(x=long_rounded, y=lat_rounded, color=log(pop)), size = gridsizeinkm * .04) +
+
+geostat_data_rastered <- geostat_data_rastered %>% group_by(utm_x_rounded, utm_y_rounded, lat_rounded, long_rounded) %>% summarise(TOT_P=sum(TOT_P))
+
+ggplot(geostat_data_rastered) + 
+    annotation_map(world_map, fill="gray50", colour="gray50") +
+    geom_point(aes(x=long_rounded, y=lat_rounded, color=log(TOT_P)), shape=15, size=cellsize) +
     scale_color_distiller(palette = "Spectral") +
@@
 ```{r per_capita_scatterplot, messages=FALSE, echo=FALSE}
-  num_vs_pop <- join_by_utm_rounded(ao_data_subset, geostat_data)
+  num_vs_pop <- join_by_utm_rounded(ao_data_subset, geostat_data_rastered)
 
-  ggplot(num_vs_pop, aes(x=pop, y=num)) + 
+  ggplot(num_vs_pop, aes(x=TOT_P, y=num)) + 
     geom_point()+
@@
 
-The number of `r summarize_list(genera)` in Norway per `r gridsizeinkm` x `r gridsizeinkm` km cs cells, versus the human population in those cells.
-R² of the linaer regression is `r summary(lm(num ~ pop, data=num_vs_pop))$r.squared`.
-R² of the log linaer regression is `r summary(lm(log10(num) ~ log10(pop), data=num_vs_pop))$r.squared`.
+The number of `r summarize_list(obs_filter[-1])` in Norway per `r gridsizeinkm` x `r gridsizeinkm` km cs cells, versus the human population in those cells.
+R² of the linaer regression is `r summary(lm(num ~ TOT_P, data=num_vs_pop))$r.squared`.
+R² of the log linaer regression is `r summary(lm(log10(num) ~ log10(TOT_P), data=num_vs_pop))$r.squared`.
 
@@
 
-num_vs_pop$percapita <- num_vs_pop$num / num_vs_pop$pop
+num_vs_pop$percapita <- num_vs_pop$num / num_vs_pop$TOT_P
 
 ggplot(num_vs_pop) + 
-    annotation_map(world_map, fill="gray") +
-    geom_point(aes(x=long_rounded, y=lat_rounded, color=log(percapita)), size = gridsizeinkm * .04) +
+    annotation_map(world_map, fill="gray50", colour="gray50") +
+    geom_point(aes(x=long_rounded, y=lat_rounded, color=log(percapita)), shape=15, size=cellsize) +
     scale_color_distiller(palette = "Spectral") +
@@
 
-The log number of `r summarize_list(genera)` observations in Norway per capita within `r gridsizeinkm` x `r gridsizeinkm` km cells.
+The log number of `r summarize_list(obs_filter[-1])` observations in Norway per capita within `r gridsizeinkm` x `r gridsizeinkm` km cells.

### 2018-02-15 exploration.R

@@ -4,30 +4,54 @@ read_gbif <- function(path) {
 }
 
 read_geostat <- function(path) {
-  result <- read.table(path, sep=",", header=TRUE, quote = "", fill=TRUE)
+  result <- read.table(path, sep=",", header=TRUE, fill=TRUE)
   result
 }
 
-geostat_add_utm <- function(dataset) {
-  dataset$utm_y <- strtoi(substr(dataset$X.GRD_ID, 6, 9)) * 1000
-  dataset$utm_x <- strtoi(substr(dataset$X.GRD_ID, 11, 14)) * 1000
+filter_observations <- function(dataset, filter) {
+  result <- subset(dataset, dataset[,filter[1]] %in% filter[-1] & taxonrank == "SPECIES")
+  result <- droplevels(result)
+  result
+}
+
+rasterize <- function(dataset, km=1) {
+  if(!("utm_x" %in% colnames(dataset))) {
+    dataset <- add_UTM(dataset)
+  }
   
-  xy <- data.frame(ID = 1:length(dataset[, 1]), X = dataset$utm_x, Y = dataset$utm_y)
+  dataset$utm_x_rounded <- (dataset$utm_x %/% (1000 * km)) * (1000 * km)
+  dataset$utm_y_rounded <- (dataset$utm_y %/% (1000 * km)) * (1000 * km)
+  
+  xy <- data.frame(ID = 1:length(dataset[, 1]), X = dataset$utm_x_rounded, Y = dataset$utm_y_rounded)
   coordinates(xy) <- c("X", "Y")
   proj4string(xy) <- CRS("+proj=laea +lat_0=52 +lon_0=10 +x_0=4321000 +y_0=3210000 +ellps=GRS80 +units=m +no_defs")
   res <- spTransform(xy, CRS("+proj=longlat +datum=WGS84"))
   res <- as.data.frame(res)
   
-  dataset$long <- res$X
-  dataset$lat <- res$Y
-
+  dataset$long_rounded <- res$X
+  dataset$lat_rounded <- res$Y
+  
   return(dataset)
 }
 
-filter_by_genus <- function(dataset, genera) {
-  result <- subset(dataset, genus %in% genera & taxonrank == "SPECIES")
-  result <- droplevels(result)
-  result
+
+add_UTM <- function(dataset) {
+  if("GRD_ID" %in% colnames(dataset)) {
[…]

### 2018-02-15 paper.Rmd
Improve legend readability

@@
 lat_dataset <- c(min(ao_data_subset$decimallatitude) %/% 1 , max(ao_data_subset$decimallatitude) %/% 1 + 1)
+
+gridsizeinkm <- 50
+cellsize <- (lat_dataset[2] - lat_dataset[1]) * gridsizeinkm * .0016
+
 ```
@@
     theme(panel.background = element_rect(fill="aliceblue"), legend.position = "right") +
+    guides(color = guide_legend(override.aes = list(size=4))) +
     coord_map(projection = "gnomonic", orientation=c(60, 16, -5),
@@
 ```{r rasterize, message=FALSE, echo=FALSE}
-  gridsizeinkm <- 50
-  cellsize <- (lat_dataset[2] - lat_dataset[1]) * gridsizeinkm * .0016
   ao_data_subset <- rasterize(ao_data_subset, gridsizeinkm)

### 2018-02-15 paper.Rmd
Add map comparing observations to linear model generated

@@
 
-gridsizeinkm <- 50
-cellsize <- (lat_dataset[2] - lat_dataset[1]) * gridsizeinkm * .0016
+gridsizeinkm <- 30
+cellsize <- (lat_dataset[2] - lat_dataset[1]) * gridsizeinkm * .0018
 
@@
 R² of the linaer regression is `r summary(lm(num ~ TOT_P, data=num_vs_pop))$r.squared`.
-R² of the log linaer regression is `r summary(lm(log10(num) ~ log10(TOT_P), data=num_vs_pop))$r.squared`.
 
@@
 The log number of `r summarize_list(obs_filter[-1])` observations in Norway per capita within `r gridsizeinkm` x `r gridsizeinkm` km cells.
+
+```{r relative_to_lm_plot_rasterized, messages=FALSE, echo=FALSE}
+
+B0 <- (summary(lm(num ~ TOT_P, data=num_vs_pop))$coefficients)["(Intercept)","Estimate"]
+B1 <- (summary(lm(num ~ TOT_P, data=num_vs_pop))$coefficients)["TOT_P","Estimate"]
+
+num_vs_pop$relativeToLM <- num_vs_pop$num / (B0 + B1 * num_vs_pop$TOT_P)
+
+ggplot(num_vs_pop) + 
+    annotation_map(world_map, fill="gray50", colour="gray50") +
+    geom_point(aes(x=long_rounded, y=lat_rounded, color=log(relativeToLM)), shape=15, size=cellsize) +
+    scale_color_distiller(palette = "Spectral") +
+    annotation_map(world_map, fill=NA, colour="gray30") +
+    theme(panel.background = element_rect(fill="aliceblue"), legend.position = "right") +
+    coord_map(projection = "gnomonic", orientation=c(60, 16, -5),
+              xlim=long_dataset,  # limits on longitude
+              ylim=lat_dataset) + # limits on latitude
+   xlab("Longitude") + ylab("Latitude")
+
+```
+
+The log number of `r summarize_list(obs_filter[-1])` observations in Norway divided by the number predicted by the regression line (numberOfObservations = `r B0` + `r B1` * humanPopulation),  within `r gridsizeinkm` x `r gridsizeinkm` km cells.

### 2018-03-07 paper.Rmd
Major revision: use spatial data frames and rasters instead of home-brew, get data from GBIF API.

--- 2018-02-15 paper.Rmd
@@
 ```{r setup, include=FALSE, echo=FALSE}
+# library(plyr)
+
+# library(rgdal)
+# library(ggalt)
+
+library(dplyr)
+library(rasterVis)
+library(rgbif)
+library(sp)
+library(raster)
 library(ggplot2)
-library(plyr)
-library(sp)
-library(rgdal)
-library(ggalt)
-library(dplyr)
 
@@
 
-obs_filter <- c("genus", "Larus")
+coordinates_norway <- c(3500000, 3500000, 5500000, 5500000)
+
 ```
@@
 ```{r read_data, include=FALSE, echo=FALSE}
-ao_data <- read_gbif(path="data/Laridae-NO-2016.csv")
-geostat_data <- read_geostat(path="data/GeoStat_NO.csv")
+taxon <- rgbif::name_suggest(q='Larus', rank='Genus')
+
+  if (!file.exists("./data/GBIF_data.zip")) {
+    download_key <- occ_download(
+      paste('taxonKey =', taxon$key[1]),
+
+      'country = NO',
+      'year = 2016',
+      'month >= 3',
+      'month <= 5',
+
+      'hasCoordinate = TRUE',
+      type = "and"
+    ) %>%
+      occ_download_meta
+  }
+
+ao_data <- load_GBIF_data(download_key)
 ```
 
-```{r filter_data, include=FALSE, echo=FALSE}
-ao_data_subset <- filter_observations(ao_data, obs_filter)  # filter observations
+```{r grid, include=FALSE, echo=FALSE}}
+gridsizeinkm <- 50
 
-ao_data_subset <- rasterize(ao_data_subset)   # round all observation positions to km
-geostat_data <- rasterize(geostat_data)       # round all population data positions to km
+grid <- get_grid(gridsizeinkm, coordinates_norway)
+grid_10km <- get_grid(10, coordinates_norway)
 
-long_dataset <- c(min(ao_data_subset$decimallongitude) %/% 1 , max(ao_data_subset$decimallongitude) %/% 1 + 1)
-lat_dataset <- c(min(ao_data_subset$decimallatitude) %/% 1 , max(ao_data_subset$decimallatitude) %/% 1 + 1)
 ```
@@
 
-Filtering on `r summarize_list(obs_filter[-1])` leaves `r nrow(ao_data_subset)` observations, plotted on the map below.
+Filtering on `r taxon['canonicalName']` observations, plotted below.
 
-```{r map, echo=FALSE, message=FALSE}
-world_map <- map_data("world", "Norway")
+```{r map_per_species, echo=FALSE, message=FALSE}
 
-ggplot(ao_data_subset) + annotation_map(world_map, fill="gray50", colour="gray50") +
-    geom_point(aes(x=decimallongitude, y=decimallatitude, color=species), alpha=0.5, size=cellsize/gridsizeinkm, shape=15) +
-    theme(panel.background = element_rect(fill="aliceblue"), legend.position = "right") +
-    coord_map(projection = "gnomonic", orientation=c(60, 16, -5),
-              xlim=long_dataset,  # limits on longitude
-              ylim=lat_dataset) + # limits on latitude
+nor_map <- map_data("world", "Norway")
+sp::coordinates(nor_map) <- ~long + lat ## make spp_dq_m into SpatialPointsDataFrame
+sp::proj4string(nor_map) <- CRS("+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs +towgs84=0,0,0") ## set CRS
+
+ggplot(as.data.frame(ao_data)) +
+    annotation_map(as.data.frame(nor_map), fill="gray70", colour="gray70") +
+    geom_point(aes(x=decimalLongitude, y=decimalLatitude, color=species), size=2, shape=20) +
+    theme(panel.background = element_rect(fill="aliceblue"), legend.position = "none") +
+    coord_map(projection = "gnomonic", orientation=c(60, 16, -5)) + # limits on latitude
     xlab("Longitude") +
-    ylab("Latitude")
+    ylab("Latitude") +
+    facet_wrap(~ species)
 
@@
 
-Map for `r summarize_list(obs_filter[-1])` in Norway
+Map for each `r taxon['canonicalName']` species in Norway
 
-```{r rasterize, message=FALSE, echo=FALSE}
-  gridsizeinkm <- 50
-  cellsize <- (lat_dataset[2] - lat_dataset[1]) * gridsizeinkm * .0016
-  ao_data_subset <- rasterize(ao_data_subset, gridsizeinkm)
+```{r rastermap_observations, message=FALSE, echo=FALSE}
+
+ao_data_rasterized <- raster::rasterize(spTransform(ao_data, CRS("+proj=laea +lat_0=52 +lon_0=10 +x_0=4321000 +y_0=3210000 +ellps=GRS80 +units=m +no_defs")), grid, fun='count')
+
+nor_map <- map_data("world", "Norway")
+sp::coordinates(nor_map) <- ~long + lat ## make spp_dq_m into SpatialPointsDataFrame
+sp::proj4string(nor_map) <- CRS("+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs +towgs84=0,0,0") ## set CRS
+nor_map <- spTransform(nor_map, CRS("+proj=laea +lat_0=52 +lon_0=10 +x_0=4321000 +y_0=3210000 +ellps=GRS80 +units=m +no_defs"))
+
+gplot(ao_data_rasterized) +
+    coord_fixed() +
+    annotation_map(as.data.frame(nor_map), fill="gray40", colour="black") +
+    geom_tile(aes(fill=log(value)), alpha=1) +
+    annotation_map(as.data.frame(nor_map), fill=NA, colour="black") +
+    theme(panel.background = element_rect(fill="aliceblue"), legend.position = "right") +
+    scale_fill_distiller(palette = "Spectral", na.value=NA) +
+    guides(color = guide_legend(override.aes = list(size=4))) +
+    # coord_map(projection = "gnomonic", orientation=c(60, 16, -5)) + # limits on latitude
+    xlab("Longitude") +
+    ylab("Latitude") + 
+    labs(title="Total observation density") +
+    labs(subtitle=taxon['canonicalName']) +
+    labs(caption = paste("Density map for all", taxon['canonicalName'], "observations in Norway, rasterized to", gridsizeinkm, "x", gridsizeinkm, "km cells."))
+
 ```
@@
 ```{r rastermap, message=FALSE, echo=FALSE}
-world_map <- map_data("world", "Norway")
 
-ggplot(ao_data_subset) + 
-    annotation_map(world_map, fill="gray50", colour="gray50") +
-    geom_point(aes(x=long_rounded, y=lat_rounded, color=species), alpha=0.3, size = .5 * cellsize) +
-    theme(panel.background = element_rect(fill="aliceblue"), legend.position = "right") +
-    coord_map(projection = "gnomonic", orientation=c(60, 16, -5),
-              xlim=long_dataset,  # limits on longitude
-              ylim=lat_dataset) + # limits on latitude
-   xlab("Longitude") + ylab("Latitude")
+# grid_poly <- rasterToPolygons(grid)
+# 
+# grid_poly_latlong <- spTransform(grid_poly, CRS("+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs +towgs84=0,0,0"))
+# 
+# ao_data_rasterized <- asterize(ao_data, grid_poly_latlong, fun='count')
+# 
+# 
+# 
+# 
+# 
+# nor_map2 <- map_data("world", "Norway")
+# sp::coordinates(nor_map2) <- ~long + lat ## make spp_dq_m into SpatialPointsDataFrame
+# sp::proj4string(nor_map2) <- CRS("+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs +towgs84=0,0,0") ## set CRS
+# 
+# gplot(ao_data_rasterized2) +
+#     coord_fixed() +
+#     annotation_map(as.data.frame(nor_map2), fill="gray40", colour="black") +
+#     geom_tile(aes(fill=log(value)), alpha=1) +
+#     annotation_map(as.data.frame(nor_map2), fill=NA, colour="black") +
+#     theme(panel.background = element_rect(fill="aliceblue"), legend.position = "right") +
+#     scale_fill_distiller(palette = "Spectral", na.value=NA) +
+#     guides(color = guide_legend(override.aes = list(size=4))) +
+#     coord_map(projection = "gnomonic", orientation=c(60, 16, -5)) + # limits on latitude
+#     xlab("Longitude") +
+#     ylab("Latitude")
 
@@
 
-Map for `r summarize_list(obs_filter[-1])` in Norway, rasterized to `r gridsizeinkm` x `r gridsizeinkm` km  cells.
+## Human population density
+```{r map_human_population, messages=FALSE, echo=FALSE}
 
-```{r densitymap_2, message=FALSE, echo=FALSE}
+geostat_data <- read.table("data/GeoStat_NO.csv", sep=",", header=TRUE, fill=TRUE)
+geostat_data$y <- strtoi(substr(geostat_data$GRD_ID, 5, 8)) * 1000 + 500
+geostat_data$x <- strtoi(substr(geostat_data$GRD_ID, 10, 13)) * 1000 + 500
 
-counts <- ddply(ao_data_subset, .(ao_data_subset$long_rounded, ao_data_subset$lat_rounded), nrow)
-names(counts) <- c("long_rounded", "lat_rounded", "freq")
+sp::coordinates(geostat_data) <- ~x + y ## make spp_dq_m into SpatialPointsDataFrame
+sp::proj4string(geostat_data) <- CRS("+proj=laea +lat_0=52 +lon_0=10 +x_0=4321000 +y_0=3210000 +ellps=GRS80 +units=m +no_defs") ## set CRS
 
-ggplot(counts) + 
-    annotation_map(world_map, fill="gray50", colour="gray50") +
-    geom_point(aes(x=long_rounded, y=lat_rounded, color=log(freq)), shape=15, size=cellsize) +
-    scale_color_distiller(palette = "Spectral") +
-    annotation_map(world_map, fill=NA, colour="gray30") +
+pop_unrasterized <- raster::rasterize(geostat_data, grid_10km, "TOT_P", fun='sum')
+
+gplot(pop_unrasterized) +
+    coord_fixed() +
+    annotation_map(as.data.frame(nor_map), fill="gray40", colour="black") +
+    geom_tile(aes(fill=log(value)), alpha=1) +
+    annotation_map(as.data.frame(nor_map), fill=NA, colour="black") +
     theme(panel.background = element_rect(fill="aliceblue"), legend.position = "right") +
-    coord_map(projection = "gnomonic", orientation=c(60, 16, -5),
-              xlim=long_dataset,  # limits on longitude
-              ylim=lat_dataset) + # limits on latitude
-   xlab("Longitude") + ylab("Latitude")
-```
-
-Map for `r summarize_list(obs_filter[-1])` in Norway, raterized to `r .018 * gridsizeinkm` x `r .01  * gridsizeinkm` degree cells.
-
-## Human population density
-
-```{r human_pop_data, messages=FALSE, echo=FALSE}
-ggplot(geostat_data) + 
-    annotation_map(world_map, fill="gray") +
-    geom_point(aes(x=long_rounded, y=lat_rounded, color=log(TOT_P)), size=cellsize/gridsizeinkm, shape=15) +
-    scale_color_distiller(palette = "Spectral") +
-    annotation_map(world_map, fill=NA, colour="gray30") +
-    theme(panel.background = element_rect(fill="aliceblue"), legend.position = "right") +
-    coord_map(projection = "gnomonic", orientation=c(60, 16, -5),
-              xlim=long_dataset,  # limits on longitude
-              ylim=lat_dataset) + # limits on latitude
-   xlab("Longitude") + ylab("Latitude")
-
+    scale_fill_distiller(palette = "Spectral", na.value=NA) +
+    guides(color = guide_legend(override.aes = list(size=4))) +
+    # coord_map(projection = "gnomonic", orientation=c(60, 16, -5)) + # limits on latitude
+    xlab("Longitude") +
+    ylab("Latitude") +
+    labs(fill = "Log population") +
+    ggtitle("Human population density")
 ```
@@
 
-```{r human_pop_rasterize, messages=FALSE, echo=FALSE}
-geostat_data_rastered <- rasterize(geostat_data, gridsizeinkm)
-```
+```{r rastermap_human_population, messages=FALSE, echo=FALSE}
 
-```{r human_pop_plot_rasterized, messages=FALSE, echo=FALSE}
+pop_rasterized <- raster::rasterize(geostat_data, grid, "TOT_P", fun='sum')
 
-geostat_data_rastered <- geostat_data_rastered %>% group_by(utm_x_rounded, utm_y_rounded, lat_rounded, long_rounded) %>% summarise(TOT_P=sum(TOT_P))
-
-ggplot(geostat_data_rastered) + 
-    annotation_map(world_map, fill="gray50", colour="gray50") +
-    geom_point(aes(x=long_rounded, y=lat_rounded, color=log(TOT_P)), shape=15, size=cellsize) +
-    scale_color_distiller(palette = "Spectral") +
-    annotation_map(world_map, fill=NA, colour="gray30") +
+gplot(pop_rasterized) +
+    coord_fixed() +
+    annotation_map(as.data.frame(nor_map), fill="gray40", colour="black") +
+    geom_tile(aes(fill=log(value)), alpha=1) +
+    annotation_map(as.data.frame(nor_map), fill=NA, colour="black") +
     theme(panel.background = element_rect(fill="aliceblue"), legend.position = "right") +
-    coord_map(projection = "gnomonic", orientation=c(60, 16, -5),
-              xlim=long_dataset,  # limits on longitude
-              ylim=lat_dataset) + # limits on latitude
-   xlab("Longitude") + ylab("Latitude")
+    scale_fill_distiller(palette = "Spectral", na.value=NA) +
+    guides(color = guide_legend(override.aes = list(size=4))) +
+    # coord_map(projection = "gnomonic", orientation=c(60, 16, -5)) + # limits on latitude
+    xlab("Longitude") +
+    ylab("Latitude") +
+    labs(fill = "Log population") +
+    ggtitle("Human population density")
 
@@
 
-```{r per_capita_scatterplot, messages=FALSE, echo=FALSE}
-  num_vs_pop <- join_by_utm_rounded(ao_data_subset, geostat_data_rastered)
+```{r scatterplot_per_capita, messages=FALSE, echo=FALSE}
 
-  ggplot(num_vs_pop, aes(x=TOT_P, y=num)) + 
+  pop_rasterized.points <- as.data.frame(rasterToPoints(pop_rasterized, spatial=TRUE))[c("layer","x","y")]
+  pop_rasterized.points <- plyr::rename(pop_rasterized.points, c("layer"="population"))
+  ao_data_rasterized.points <- as.data.frame(rasterToPoints(ao_data_rasterized, spatial=TRUE))[c("ID","x","y")]
+  ao_data_rasterized.points <- plyr::rename(ao_data_rasterized.points, c("ID"="observations"))
+
+  num_vs_pop <- ao_data_rasterized.points %>% inner_join(pop_rasterized.points, by = c("x", "y"))
+
+  ggplot(num_vs_pop, aes(x=population, y=observations)) +
     geom_point()+
@@
 
-The number of `r summarize_list(obs_filter[-1])` in Norway per `r gridsizeinkm` x `r gridsizeinkm` km cs cells, versus the human population in those cells.
-R² of the linaer regression is `r summary(lm(num ~ TOT_P, data=num_vs_pop))$r.squared`.
-R² of the log linaer regression is `r summary(lm(log10(num) ~ log10(TOT_P), data=num_vs_pop))$r.squared`.
+The number of `r taxon['canonicalName']` in Norway per `r gridsizeinkm` x `r gridsizeinkm` km cs cells, versus the human population in those cells.
+R² of the linaer regression is `r summary(lm(observations ~ population, data=num_vs_pop))$r.squared`.
 
-```{r per_capita_plot_rasterized, messages=FALSE, echo=FALSE}
+```{r rastermap_observations_per_capita, messages=FALSE, echo=FALSE}
 
-num_vs_pop$percapita <- num_vs_pop$num / num_vs_pop$TOT_P
+num_vs_pop$percapita <- num_vs_pop$observations / num_vs_pop$population
 
-ggplot(num_vs_pop) + 
-    annotation_map(world_map, fill="gray50", colour="gray50") +
-    geom_point(aes(x=long_rounded, y=lat_rounded, color=log(percapita)), shape=15, size=cellsize) +
-    scale_color_distiller(palette = "Spectral") +
-    annotation_map(world_map, fill=NA, colour="gray30") +
+sp::coordinates(num_vs_pop) <- ~x + y ## make spp_dq_m into SpatialPointsDataFrame
+sp::proj4string(num_vs_pop) <- CRS("+proj=laea +lat_0=52 +lon_0=10 +x_0=4321000 +y_0=3210000 +ellps=GRS80 +units=m +no_defs") ## set CRS
+
+num_vs_pop_rasterized <- raster::rasterize(num_vs_pop, grid, "percapita", fun='sum')
+
+gplot(num_vs_pop_rasterized) +
+    coord_fixed() +
+    annotation_map(as.data.frame(nor_map), fill="gray40", colour="black") +
+    geom_tile(aes(fill=log(value)), alpha=1) +
+    annotation_map(as.data.frame(nor_map), fill=NA, colour="black") +
     theme(panel.background = element_rect(fill="aliceblue"), legend.position = "right") +
-    coord_map(projection = "gnomonic", orientation=c(60, 16, -5),
-              xlim=long_dataset,  # limits on longitude
-              ylim=lat_dataset) + # limits on latitude
-   xlab("Longitude") + ylab("Latitude")
+    scale_fill_distiller(palette = "Spectral", na.value=NA) +
+    guides(color = guide_legend(override.aes = list(size=4))) +
+    # coord_map(projection = "gnomonic", orientation=c(60, 16, -5)) + # limits on latitude
+    xlab("Longitude") +
+    ylab("Latitude") +
+    labs(fill = "Log observations") +
+    ggtitle("Observations per capita")
+
+# ggplot(num_vs_pop) + 
+#     coord_fixed() +
+#     annotation_map(as.data.frame(nor_map), fill="gray50", colour="gray50") +
+#     geom_point(aes(x=x, y=y, color=log(percapita)), shape=15) +
+#     scale_color_distiller(palette = "Spectral") +
+#     annotation_map(as.data.frame(nor_map), fill=NA, colour="gray30") +
+#     theme(panel.background = element_rect(fill="aliceblue"), legend.position = "right") +
+    # coord_map(projection = "gnomonic", orientation=c(60, 16, -5),
+    #           xlim=long_dataset,  # limits on longitude
+    #           ylim=lat_dataset) + # limits on latitude
+   # xlab("Longitude") + ylab("Latitude")
 
@@
 
-The log number of `r summarize_list(obs_filter[-1])` observations in Norway per capita within `r gridsizeinkm` x `r gridsizeinkm` km cells.
+The log number of `r taxon['canonicalName']` observations in Norway per capita within `r gridsizeinkm` x `r gridsizeinkm` km cells.
+
+```{r rastermap_observations_relative_to_lm, messages=FALSE, echo=FALSE}
+
+num_vs_pop <- as.data.frame(num_vs_pop)
+
+B0 <- (summary(lm(observations ~ population, data=num_vs_pop))$coefficients)["(Intercept)","Estimate"]
+B1 <- (summary(lm(observations ~ population, data=num_vs_pop))$coefficients)["population","Estimate"]
+
+num_vs_pop$relativeToLM <- num_vs_pop$observations / (B0 + B1 * num_vs_pop$population)
+
+sp::coordinates(num_vs_pop) <- ~x + y ## make spp_dq_m into SpatialPointsDataFrame
+sp::proj4string(num_vs_pop) <- CRS("+proj=laea +lat_0=52 +lon_0=10 +x_0=4321000 +y_0=3210000 +ellps=GRS80 +units=m +no_defs") ## set CRS
+
+num_vs_pop_rasterized_LM <- raster::rasterize(num_vs_pop, grid, "relativeToLM", fun='sum')
+
+gplot(num_vs_pop_rasterized_LM) +
+    coord_fixed() +
+    annotation_map(as.data.frame(nor_map), fill="gray40", colour="black") +
+    geom_tile(aes(fill=log(value)), alpha=1) +
+    annotation_map(as.data.frame(nor_map), fill=NA, colour="black") +
+    theme(panel.background = element_rect(fill="aliceblue"), legend.position = "right") +
+    scale_fill_distiller(palette = "Spectral", na.value=NA) +
+    guides(color = guide_legend(override.aes = list(size=4))) +
+    # coord_map(projection = "gnomonic", orientation=c(60, 16, -5)) + # limits on latitude
+    xlab("Longitude") +
+    ylab("Latitude") +
+    labs(fill = "Log observations") +
+    ggtitle("Observations relative to regression with human population")
+
+# ggplot(num_vs_pop) + 
+#     annotation_map(world_map, fill="gray50", colour="gray50") +
+#     geom_point(aes(x=long_rounded, y=lat_rounded, color=log(relativeToLM)), shape=15, size=cellsize) +
+#     scale_color_distiller(palette = "Spectral") +
+#     annotation_map(world_map, fill=NA, colour="gray30") +
+#     theme(panel.background = element_rect(fill="aliceblue"), legend.position = "right") +
+#     coord_map(projection = "gnomonic", orientation=c(60, 16, -5),
+#               xlim=long_dataset,  # limits on longitude
+#               ylim=lat_dataset) + # limits on latitude
+#    xlab("Longitude") + ylab("Latitude")
+
+```
+
+The log number of `r taxon['canonicalName']` observations in Norway divided by the number predicted by the regression line (numberOfObservations = `r B0` + `r B1` * humanPopulation),  within `r gridsizeinkm` x `r gridsizeinkm` km cells.

### 2018-03-07 exploration.R

@@ -1,75 +1,58 @@
-read_gbif <- function(path) {
-  result <- read.table(path, sep="\t", header=TRUE, fill=TRUE)
-  result
-}
-
-read_geostat <- function(path) {
-  result <- read.table(path, sep=",", header=TRUE, fill=TRUE)
-  result
-}
-
-filter_observations <- function(dataset, filter) {
-  result <- subset(dataset, dataset[,filter[1]] %in% filter[-1] & taxonrank == "SPECIES")
-  result <- droplevels(result)
-  result
-}
-
-rasterize <- function(dataset, km=1) {
-  if(!("utm_x" %in% colnames(dataset))) {
-    dataset <- add_UTM(dataset)
+download_GBIF_API <- function(download_key,n_try,Sys.sleep_duration,destfile_name){
+  start_time <- Sys.time()
+  n_try_count <- 1
+  
+  download_url <- paste("http://api.gbif.org/v1/occurrence/download/request/",
+                        download_key[1],sep="")
+  try_download <- try(download.file(url=download_url,destfile=destfile_name,
+                                    quiet=TRUE, mode="wb"),silent = TRUE)
+  
+  while (inherits(try_download, "try-error") & n_try_count < n_try) {   
+    Sys.sleep(Sys.sleep_duration)
+    n_try_count <- n_try_count+1
+    try_download <- try(download.file(url=download_url,destfile=destfile_name,
+                                      quiet=TRUE),silent = TRUE)
+    print(paste("trying... Download link not ready. Time elapsed (min):",
+                round(as.numeric(paste(difftime(Sys.time(),start_time, units = "mins"))),2)))
   }
   
-  dataset$utm_x_rounded <- (dataset$utm_x %/% (1000 * km)) * (1000 * km)
-  dataset$utm_y_rounded <- (dataset$utm_y %/% (1000 * km)) * (1000 * km)
-  
-  xy <- data.frame(ID = 1:length(dataset[, 1]), X = dataset$utm_x_rounded, Y = dataset$utm_y_rounded)
-  coordinates(xy) <- c("X", "Y")
-  proj4string(xy) <- CRS("+proj=laea +lat_0=52 +lon_0=10 +x_0=4321000 +y_0=3210000 +ellps=GRS80 +units=m +no_defs")
[…]

### 2018-03-07 paper.Rmd
Fix Typo
Improved download of GBIF and population data

@@
 ```{r setup, include=FALSE, echo=FALSE}
-# library(plyr)
-
-# library(rgdal)
-# library(ggalt)
-
   library(dplyr)
@@
 
-coordinates_norway <- c(3500000, 3500000, 5500000, 5500000)
+  country_code <- "NO"
+  country_coordinates <- c(3500000, 3500000, 5500000, 5500000)
+  country_map <- map_data("world", country_code)
 
+  ## Season onsets for Værnes airport according to Norwegian Meteorological Institute
+  spring <- as.Date("3-15",  format = "%m-%d") ## From Mar 15 
+  summer <- as.Date("5-22",  format = "%m-%d")  ## From May 22
+  fall <- as.Date("9-11",  format = "%m-%d")  ## From Sep 11
+  winter <- as.Date("11-18",  format = "%m-%d")  ## From Nov 18
+
+  sp::coordinates(country_map) <- ~long + lat
+  sp::proj4string(country_map) <- CRS("+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs +towgs84=0,0,0")
 ```
@@
 
-  if (!file.exists("./data/GBIF_data.zip")) {
-    download_key <- occ_download(
+  gbif_filter <- list(
       paste('taxonKey =', taxon$key[1]),
+      paste('country =', country_code),
+      'datasetKey = b124e1e0-4755-430f-9eab-894f25a9b59c',
+      'year = 2016',
+      'hasCoordinate = TRUE',
+      'hasGeospatialIssue = FALSE',
+      type = "and")
 
-      'country = NO',
-      'year = 2016',
-      'month >= 3',
-      'month <= 5',
+  ao_data <- load_GBIF_data(gbif_filter)
+  geostat_data <- load_GeoStat_data(country_code)
 
-      'hasCoordinate = TRUE',
-      type = "and"
-    ) %>%
-      occ_download_meta
-  }
-
-ao_data <- load_GBIF_data(download_key)
 ```
 
-```{r grid, include=FALSE, echo=FALSE}}
+```{r grid, include=FALSE, echo=FALSE}
   gridsizeinkm <- 50
 
-grid <- get_grid(gridsizeinkm, coordinates_norway)
-grid_10km <- get_grid(10, coordinates_norway)
+  grid <- get_grid(gridsizeinkm, country_coordinates)
+  grid_10km <- get_grid(10, country_coordinates)
 
@@
 ```{r map_per_species, echo=FALSE, message=FALSE}
-
-nor_map <- map_data("world", "Norway")
-sp::coordinates(nor_map) <- ~long + lat ## make spp_dq_m into SpatialPointsDataFrame
-sp::proj4string(nor_map) <- CRS("+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs +towgs84=0,0,0") ## set CRS
-
   ggplot(as.data.frame(ao_data)) +
-    annotation_map(as.data.frame(nor_map), fill="gray70", colour="gray70") +
-    geom_point(aes(x=decimalLongitude, y=decimalLatitude, color=species), size=2, shape=20) +
+    annotation_map(as.data.frame(country_map), fill="gray70", colour="gray70") +
+    geom_point(aes(x=decimalLongitude, y=decimalLatitude, color=species), size=.5, shape=20) +
     theme(panel.background = element_rect(fill="aliceblue"), legend.position = "none") +
-    coord_map(projection = "gnomonic", orientation=c(60, 16, -5)) + # limits on latitude
+    coord_map(projection = "gnomonic", orientation=c(60, 16, -5)) +
     xlab("Longitude") +
@@
 
-nor_map <- map_data("world", "Norway")
-sp::coordinates(nor_map) <- ~long + lat ## make spp_dq_m into SpatialPointsDataFrame
-sp::proj4string(nor_map) <- CRS("+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs +towgs84=0,0,0") ## set CRS
-nor_map <- spTransform(nor_map, CRS("+proj=laea +lat_0=52 +lon_0=10 +x_0=4321000 +y_0=3210000 +ellps=GRS80 +units=m +no_defs"))
+  country_map <- spTransform(country_map, CRS("+proj=laea +lat_0=52 +lon_0=10 +x_0=4321000 +y_0=3210000 +ellps=GRS80 +units=m +no_defs"))
 
@@
     coord_fixed() +
-    annotation_map(as.data.frame(nor_map), fill="gray40", colour="black") +
+    annotation_map(as.data.frame(country_map), fill="gray40", colour="black") +
     geom_tile(aes(fill=log(value)), alpha=1) +
-    annotation_map(as.data.frame(nor_map), fill=NA, colour="black") +
+    annotation_map(as.data.frame(country_map), fill=NA, colour="black") +
     theme(panel.background = element_rect(fill="aliceblue"), legend.position = "right") +
@@
     guides(color = guide_legend(override.aes = list(size=4))) +
-    # coord_map(projection = "gnomonic", orientation=c(60, 16, -5)) + # limits on latitude
+    # coord_map(projection = "gnomonic", orientation=c(60, 16, -5)) +
     xlab("Longitude") +
@@
 
-```{r rastermap, message=FALSE, echo=FALSE}
-
-# grid_poly <- rasterToPolygons(grid)
-# 
-# grid_poly_latlong <- spTransform(grid_poly, CRS("+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs +towgs84=0,0,0"))
-# 
-# ao_data_rasterized <- asterize(ao_data, grid_poly_latlong, fun='count')
-# 
-# 
-# 
-# 
-# 
-# nor_map2 <- map_data("world", "Norway")
-# sp::coordinates(nor_map2) <- ~long + lat ## make spp_dq_m into SpatialPointsDataFrame
-# sp::proj4string(nor_map2) <- CRS("+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs +towgs84=0,0,0") ## set CRS
-# 
-# gplot(ao_data_rasterized2) +
-#     coord_fixed() +
-#     annotation_map(as.data.frame(nor_map2), fill="gray40", colour="black") +
-#     geom_tile(aes(fill=log(value)), alpha=1) +
-#     annotation_map(as.data.frame(nor_map2), fill=NA, colour="black") +
-#     theme(panel.background = element_rect(fill="aliceblue"), legend.position = "right") +
-#     scale_fill_distiller(palette = "Spectral", na.value=NA) +
-#     guides(color = guide_legend(override.aes = list(size=4))) +
-#     coord_map(projection = "gnomonic", orientation=c(60, 16, -5)) + # limits on latitude
-#     xlab("Longitude") +
-#     ylab("Latitude")
-
-```
-
 ## Human population density
 ```{r map_human_population, messages=FALSE, echo=FALSE}
-
-geostat_data <- read.table("data/GeoStat_NO.csv", sep=",", header=TRUE, fill=TRUE)
-geostat_data$y <- strtoi(substr(geostat_data$GRD_ID, 5, 8)) * 1000 + 500
-geostat_data$x <- strtoi(substr(geostat_data$GRD_ID, 10, 13)) * 1000 + 500
-
-sp::coordinates(geostat_data) <- ~x + y ## make spp_dq_m into SpatialPointsDataFrame
-sp::proj4string(geostat_data) <- CRS("+proj=laea +lat_0=52 +lon_0=10 +x_0=4321000 +y_0=3210000 +ellps=GRS80 +units=m +no_defs") ## set CRS
-
   pop_unrasterized <- raster::rasterize(geostat_data, grid_10km, "TOT_P", fun='sum')
@@
 
-  ggplot(num_vs_pop, aes(x=population, y=observations)) +
+  # num_vs_pop <- subset(num_vs_pop, population<100000 & observations < 1000)
+
+  sp::coordinates(num_vs_pop) <- ~x + y
+  sp::proj4string(num_vs_pop) <- CRS("+proj=laea +lat_0=52 +lon_0=10 +x_0=4321000 +y_0=3210000 +ellps=GRS80 +units=m +no_defs")
+
+  ggplot(as.data.frame(num_vs_pop), aes(x=population, y=observations)) +
     geom_point()+
@@
 num_vs_pop$percapita <- num_vs_pop$observations / num_vs_pop$population
-
-sp::coordinates(num_vs_pop) <- ~x + y ## make spp_dq_m into SpatialPointsDataFrame
-sp::proj4string(num_vs_pop) <- CRS("+proj=laea +lat_0=52 +lon_0=10 +x_0=4321000 +y_0=3210000 +ellps=GRS80 +units=m +no_defs") ## set CRS
-
 num_vs_pop_rasterized <- raster::rasterize(num_vs_pop, grid, "percapita", fun='sum')
@@
     ggtitle("Observations per capita")
-
-# ggplot(num_vs_pop) + 
-#     coord_fixed() +
-#     annotation_map(as.data.frame(nor_map), fill="gray50", colour="gray50") +
-#     geom_point(aes(x=x, y=y, color=log(percapita)), shape=15) +
-#     scale_color_distiller(palette = "Spectral") +
-#     annotation_map(as.data.frame(nor_map), fill=NA, colour="gray30") +
-#     theme(panel.background = element_rect(fill="aliceblue"), legend.position = "right") +
-    # coord_map(projection = "gnomonic", orientation=c(60, 16, -5),
-    #           xlim=long_dataset,  # limits on longitude
-    #           ylim=lat_dataset) + # limits on latitude
-   # xlab("Longitude") + ylab("Latitude")
-
 ```
@@
 
-sp::coordinates(num_vs_pop) <- ~x + y ## make spp_dq_m into SpatialPointsDataFrame
-sp::proj4string(num_vs_pop) <- CRS("+proj=laea +lat_0=52 +lon_0=10 +x_0=4321000 +y_0=3210000 +ellps=GRS80 +units=m +no_defs") ## set CRS
+sp::coordinates(num_vs_pop) <- ~x + y
+sp::proj4string(num_vs_pop) <- CRS("+proj=laea +lat_0=52 +lon_0=10 +x_0=4321000 +y_0=3210000 +ellps=GRS80 +units=m +no_defs")
 
@@
     ggtitle("Observations relative to regression with human population")
-
-# ggplot(num_vs_pop) + 
-#     annotation_map(world_map, fill="gray50", colour="gray50") +
-#     geom_point(aes(x=long_rounded, y=lat_rounded, color=log(relativeToLM)), shape=15, size=cellsize) +
-#     scale_color_distiller(palette = "Spectral") +
-#     annotation_map(world_map, fill=NA, colour="gray30") +
-#     theme(panel.background = element_rect(fill="aliceblue"), legend.position = "right") +
-#     coord_map(projection = "gnomonic", orientation=c(60, 16, -5),
-#               xlim=long_dataset,  # limits on longitude
-#               ylim=lat_dataset) + # limits on latitude
-#    xlab("Longitude") + ylab("Latitude")
-
 ```

### 2018-03-07 exploration.R

@@ -1,3 +1,9 @@
+get_GBIF_credentials <- function() {
+    options(gbif_user=rstudioapi::showPrompt(title = "GBIF username", message = "GBIF username", default = ""))
+    options(gbif_email=rstudioapi::showPrompt(title = "GBIF e-mail address", message = "GBIF e-mail address", default = ""))
+    options(gbif_pwd=rstudioapi::askForPassword("GBIF password"))
+}
+
 download_GBIF_API <- function(download_key,n_try,Sys.sleep_duration,destfile_name){
   start_time <- Sys.time()
   n_try_count <- 1
@@ -24,35 +30,73 @@ download_GBIF_API <- function(download_key,n_try,Sys.sleep_duration,destfile_nam
   }
 }
 
-load_GBIF_data <- function(download_key) {
+load_GBIF_data <- function(download_filter) {
   
-  destfile="./data/GBIF_data.zip"
+  destfile = "./data/GBIF_data.zip"
+  filter_file = "./data/GBIF_filter.txt"
+  file_age = as.numeric((difftime(as.Date(file.info(destfile)$ctime), Sys.Date(), units = "days")))
   
-  # If the data has not been downloaded already, ask for credentioals and do so now
-  if (!file.exists(destfile)) {
-    options(gbif_user=rstudioapi::showPrompt(title = "GBIF username", message = "GBIF username", default = ""))
-    options(gbif_email=rstudioapi::showPrompt(title = "GBIF e-mail address", message = "GBIF e-mail address", default = ""))
-    options(gbif_pwd=rstudioapi::askForPassword("GBIF password"))
+  
+  # If the GBIF data has been deleted, is older than 10 days or the filter has been changed, download new data
+  if(!file.exists(destfile) || file_age > 10 || paste(download_filter, collapse=', ') != paste(readChar(filter_file, file.info(filter_file)$size))) {
+    cat(paste(gbif_filter, collapse=', '), file=filter_file, sep="")
     
+    # Get the credentials if needed
+    if(!(exists("gbif_user") && exists("gbif_email") && exists("gbif_pwd"))) {
+      get_GBIF_credentials()
+    }
+    
[…]

### 2018-04-16 absences.Rmd
Generate pseudoabsences

---
title: "Norwegian citizen science data - absence data"
author: "Wouter Koch, NBIC"

output:
  pdf_document: default
  word_document: default
  html_document: default
---

```{r setup, include=FALSE, echo=FALSE}
  library(dplyr)
  library(rasterVis)
  library(rgbif)
  library(sp)
  library(raster)
  library(ggplot2)

  source("exploration.R")

  country_code <- "NO"
  country_coordinates <- c(3500000, 3500000, 5500000, 5500000)
  country_map <- map_data("world", country_code)

  ## Season onsets for Værnes airport according to Norwegian Meteorological Institute
  spring <- as.Date("2012-03-15",  format = "%Y-%m-%d") ## From Mar 15 
  summer <- as.Date("2012-05-22",  format = "%Y-%m-%d")  ## From May 22
  fall <- as.Date("2012-09-11",  format = "%Y-%m-%d")  ## From Sep 11
  winter <- as.Date("2012-11-18",  format = "%Y-%m-%d")  ## From Nov 18

  sp::coordinates(country_map) <- ~long + lat
  sp::proj4string(country_map) <- CRS("+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs +towgs84=0,0,0")
```

******

```{r variables, include=FALSE, echo=FALSE}
  higherLevel <- "Family"
  focalSpecies <- "Haliaeetus albicilla"

  gridsizeinkm <- 50
  grid <- get_grid(gridsizeinkm, country_coordinates)

```

```{r read_data, include=FALSE, echo=FALSE}
  higherLevelName <- name_backbone(name=focalSpecies, rank='species')[tolower(higherLevel)][[1]]
  taxon <- rgbif::name_suggest(q=higherLevelName, rank=higherLevel)

  gbif_filter <- list(
      paste('taxonKey =', taxon$key[1]),
      paste('country =', country_code),
      'datasetKey = b124e1e0-4755-430f-9eab-894f25a9b59c',
      'year = 2016',
      'hasCoordinate = TRUE',
      'hasGeospatialIssue = FALSE',
      type = "and")

  ao_data <- load_GBIF_data(gbif_filter)
  # geostat_data <- load_GeoStat_data(country_code)

```

```{r split_obs, include=FALSE, echo=FALSE}
  library(tidyr)

  ao_data <- subset(as.data.frame(ao_data), taxonRank == "SPECIES")
  ao_data <- ao_data %>% tidyr::separate_rows(recordedBy, sep = ",")
  ao_data$recordedBy <- as.factor(ao_data$recordedBy)

  per_user <- ao_data %>% group_by(recordedBy, species) %>% tally %>% tidyr::spread(species, n, fill = 0)
  per_user$observations <- (ao_data %>% group_by(recordedBy) %>% tally)$n

  per_user$species <- rowSums(per_user[c(-1)] > 0) - 1
  per_user <- per_user[c("recordedBy",focalSpecies,"observations","species")]

  names(per_user) <- c("recordedBy","observationsOfSpecies","totalObservations","numberOfSpeciesObserved")

```

```{r get_observers, include=FALSE, echo=FALSE}
  # Get all people that have reported this species, and at least x observations and y species
  per_user <- per_user %>% filter(observationsOfSpecies > 0 & totalObservations > 100 & numberOfSpeciesObserved > 3)

```

```{r get_empty_squares, include=FALSE, echo=FALSE}
  # For this species: get cells where it has been observed

  focus_species_observations <- ao_data %>% filter(species == focalSpecies)

  sp::coordinates(focus_species_observations) <- ~decimalLongitude + decimalLatitude
  sp::proj4string(focus_species_observations) <- CRS("+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs +towgs84=0,0,0")
  focus_species_observations <- spTransform(focus_species_observations, CRS("+proj=laea +lat_0=52 +lon_0=10 +x_0=4321000 +y_0=3210000 +ellps=GRS80 +units=m +no_defs"))
   focus_species_cells <- as.data.frame(focus_species_observations)
  focus_species_cells$decimalLatitude <- focus_species_cells$decimalLatitude %/% (gridsizeinkm * 1000) * (gridsizeinkm * 1000) + (gridsizeinkm * 500)
  focus_species_cells$decimalLongitude <- focus_species_cells$decimalLongitude %/% (gridsizeinkm * 1000) * (gridsizeinkm * 1000) + (gridsizeinkm * 500)

  focus_species_cells$cell <- paste(focus_species_cells$decimalLatitude, focus_species_cells$decimalLongitude, sep = "_")

  focus_species_cells <- focus_species_cells[c("cell")]
  focus_species_cells <- unique(focus_species_cells)

```

```{r get_absence, include=FALSE, echo=FALSE}
  # For such cells, find the number of filtered people that have been there
  # Or the number of species reported there by such filtered people

  other_species_observations <- ao_data %>% filter(species != focalSpecies)
  other_species_observations$date <- paste(other_species_observations$year, other_species_observations$month, other_species_observations$day, sep = "-")

  sp::coordinates(other_species_observations) <- ~decimalLongitude + decimalLatitude
  sp::proj4string(other_species_observations) <- CRS("+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs +towgs84=0,0,0")
  other_species_observations <- spTransform(other_species_observations, CRS("+proj=laea +lat_0=52 +lon_0=10 +x_0=4321000 +y_0=3210000 +ellps=GRS80 +units=m +no_defs"))

  other_species_observations <- as.data.frame(other_species_observations)
  other_species_observations$decimalLatitude <- other_species_observations$decimalLatitude %/% (gridsizeinkm * 1000) * (gridsizeinkm * 1000) + (gridsizeinkm * 500)
  other_species_observations$decimalLongitude <- other_species_observations$decimalLongitude %/% (gridsizeinkm * 1000) * (gridsizeinkm * 1000) + (gridsizeinkm * 500)

  other_species_observations <- other_species_observations[c("recordedBy","decimalLatitude","decimalLongitude","date")]
  other_species_observations <- unique(other_species_observations)

  other_species_observations$cell <- paste(other_species_observations$decimalLatitude, other_species_observations$decimalLongitude, sep = "_")

  other_species_observations <- subset(other_species_observations, recordedBy %in% per_user$recordedBy)
  other_species_observations <- subset(other_species_observations, !(cell %in% focus_species_cells$cell))

  sp::coordinates(other_species_observations) <- ~decimalLongitude + decimalLatitude
  sp::proj4string(other_species_observations) <- CRS("+proj=laea +lat_0=52 +lon_0=10 +x_0=4321000 +y_0=3210000 +ellps=GRS80 +units=m +no_defs")

  other_species_rasterized <- raster::rasterize(spTransform(other_species_observations, CRS("+proj=laea +lat_0=52 +lon_0=10 +x_0=4321000 +y_0=3210000 +ellps=GRS80 +units=m +no_defs")), grid, fun='count')

  country_map <- spTransform(country_map, CRS("+proj=laea +lat_0=52 +lon_0=10 +x_0=4321000 +y_0=3210000 +ellps=GRS80 +units=m +no_defs"))

```

```{r plots_presence, message=FALSE, echo=FALSE}

  focus_species_rasterized <- raster::rasterize(spTransform(focus_species_observations, CRS("+proj=laea +lat_0=52 +lon_0=10 +x_0=4321000 +y_0=3210000 +ellps=GRS80 +units=m +no_defs")), grid, fun='count')

  gplot(focus_species_rasterized) +
    coord_fixed() +
    annotation_map(as.data.frame(country_map), fill="grey50", colour="black") +
    geom_tile(aes(fill=value), alpha=1) +
    annotation_map(as.data.frame(country_map), fill=NA, colour="black") +
    theme(panel.background = element_rect(fill="aliceblue"), legend.position = "right") +
    # scale_fill_continuous(high = "red", low = "white", na.value=NA) +
    guides(color = guide_legend(override.aes = list(size=4))) +
    # coord_map(projection = "gnomonic", orientation=c(60, 16, -5)) +
    xlab("Longitude") +
    ylab("Latitude") +
    labs(fill = "Observations") + 
    ggtitle("Presence")
```

Observations of _`r focalSpecies`_ in Norway.

```{r plots_absence, message=FALSE, echo=FALSE}

  gplot(other_species_rasterized) +
    coord_fixed() +
    annotation_map(as.data.frame(country_map), fill="grey50", colour="black") +
    geom_tile(aes(fill=value), alpha=1) +
    annotation_map(as.data.frame(country_map), fill=NA, colour="black") +
    theme(panel.background = element_rect(fill="aliceblue"), legend.position = "right") +
    # scale_fill_continuous(high = "red", low = "white", na.value=NA) +
    guides(color = guide_legend(override.aes = list(size=4))) +
    # coord_map(projection = "gnomonic", orientation=c(60, 16, -5)) +
    xlab("Longitude") +
    ylab("Latitude") +
    labs(fill = "Absences") + 
    ggtitle("Absence")

```

Absence of _`r focalSpecies`_ in Norway.
(Visits by users known to report _`r focalSpecies`_, where that user reported at least one observation of other species within the same `r tolower(higherLevel)` within that cell.)

### 2018-04-16 paper.Rmd
Generate pseudoabsences

@@
   ## Season onsets for Værnes airport according to Norwegian Meteorological Institute
-  spring <- as.Date("3-15",  format = "%m-%d") ## From Mar 15 
-  summer <- as.Date("5-22",  format = "%m-%d")  ## From May 22
-  fall <- as.Date("9-11",  format = "%m-%d")  ## From Sep 11
-  winter <- as.Date("11-18",  format = "%m-%d")  ## From Nov 18
+  spring <- as.Date("2012-03-15",  format = "%Y-%m-%d") ## From Mar 15 
+  summer <- as.Date("2012-05-22",  format = "%Y-%m-%d")  ## From May 22
+  fall <- as.Date("2012-09-11",  format = "%Y-%m-%d")  ## From Sep 11
+  winter <- as.Date("2012-11-18",  format = "%Y-%m-%d")  ## From Nov 18
 
@@
 
-A test data exploration of some [Norwegian Species Observation Service](https://www.artsobservasjoner.no) data, obtained through [GBIF](https://www.gbif.org) (N = `r nrow(ao_data)`).
+A test data exploration of some [Norwegian Species Observation Service](https://www.artsobservasjoner.no) data: observations within the `r tolower(taxon['rank'])` `r taxon['canonicalName']`, obtained through [GBIF](https://www.gbif.org) (N = `r nrow(ao_data)`).
 
-Filtering on `r taxon['canonicalName']` observations, plotted below.
+## Observations
 
 ```{r map_per_species, echo=FALSE, message=FALSE}
-  ggplot(as.data.frame(ao_data)) +
-    annotation_map(as.data.frame(country_map), fill="gray70", colour="gray70") +
-    geom_point(aes(x=decimalLongitude, y=decimalLatitude, color=species), size=.5, shape=20) +
+  ggplot(subset(as.data.frame(ao_data),specificEpithet!="")) +
+    annotation_map(as.data.frame(country_map), fill="white", colour="white") +
+    geom_point(aes(x=decimalLongitude, y=decimalLatitude), color="darkgreen", alpha=.05, size=.25, shape=20) +
     theme(panel.background = element_rect(fill="aliceblue"), legend.position = "none") +
@@
     ylab("Latitude") +
-    facet_wrap(~ species)
+    facet_wrap(~ species) + 
+    ggtitle("Observations per species")
 
@@
 
-Map for each `r taxon['canonicalName']` species in Norway
+Map of observations of each species within the `r tolower(taxon['rank'])` `r taxon['canonicalName']` in Norway.
+
+```{r map_per_season, echo=FALSE, message=FALSE}
+  ggplot(as.data.frame(ao_data)) +
+    annotation_map(as.data.frame(country_map), fill="white", colour="white") +
+    geom_point(aes(x=decimalLongitude, y=decimalLatitude), color="darkgreen", alpha=.05, size=.25, shape=20) +
+    theme(panel.background = element_rect(fill="aliceblue"), legend.position = "none") +
+    coord_map(projection = "gnomonic", orientation=c(60, 16, -5)) +
+    xlab("Longitude") +
+    ylab("Latitude") +
+    facet_wrap(~ season) + 
+    ggtitle("Observations per season")
+
+```
+
+Map of all observations within the `r tolower(taxon['rank'])` `r taxon['canonicalName']` in Norway, per season.
 
@@
     coord_fixed() +
-    annotation_map(as.data.frame(country_map), fill="gray40", colour="black") +
+    annotation_map(as.data.frame(country_map), fill="grey50", colour="black") +
     geom_tile(aes(fill=log(value)), alpha=1) +
@@
     theme(panel.background = element_rect(fill="aliceblue"), legend.position = "right") +
-    scale_fill_distiller(palette = "Spectral", na.value=NA) +
+    # scale_fill_continuous(high = "red", low = "white", na.value=NA) +
+    scale_fill_distiller(palette = "Spectral", na.value=NA, direction = -1) +
     guides(color = guide_legend(override.aes = list(size=4))) +
@@
     ylab("Latitude") +
-    labs(title="Total observation density") +
-    labs(subtitle=taxon['canonicalName']) +
-    labs(caption = paste("Density map for all", taxon['canonicalName'], "observations in Norway, rasterized to", gridsizeinkm, "x", gridsizeinkm, "km cells."))
+    labs(fill = "Observations (log)") + 
+    ggtitle("Observation density")
 
@@
 
-## Human population density
+Density map for all observations in Norway within the `r tolower(taxon['rank'])` `r taxon['canonicalName']`, rasterized to `r gridsizeinkm` x `r gridsizeinkm` km cells.
+
+```{r rastermap_observations_seasonal, message=FALSE, echo=FALSE}
+
+  # ao_data_rasterized_winter <- raster::rasterize(spTransform(subset(ao_data, season=='Winter'), CRS("+proj=laea +lat_0=52 +lon_0=10 +x_0=4321000 +y_0=3210000 +ellps=GRS80 +units=m +no_defs")), grid, fun='count', field='season')
+  # ao_data_rasterized_spring <- raster::rasterize(spTransform(subset(ao_data, season=='Spring'), CRS("+proj=laea +lat_0=52 +lon_0=10 +x_0=4321000 +y_0=3210000 +ellps=GRS80 +units=m +no_defs")), grid, fun='count', field='season')
+  # ao_data_rasterized_summer <- raster::rasterize(spTransform(subset(ao_data, season=='Summer'), CRS("+proj=laea +lat_0=52 +lon_0=10 +x_0=4321000 +y_0=3210000 +ellps=GRS80 +units=m +no_defs")), grid, fun='count', field='season')
+  # ao_data_rasterized_fall <- raster::rasterize(spTransform(subset(ao_data, season=='Fall'), CRS("+proj=laea +lat_0=52 +lon_0=10 +x_0=4321000 +y_0=3210000 +ellps=GRS80 +units=m +no_defs")), grid, fun='count', field='season')
+
+  # ao_season_stack <- stack(ao_data_rasterized_winter, ao_data_rasterized_spring, ao_data_rasterized_summer, ao_data_rasterized_fall)
+  # ao_season_stack <- log(ao_season_stack)
+  # names(ao_season_stack) <- c('Winter','Spring','Summer','Fall')
+  # 
+  # levelplot(ao_season_stack,
+  #     col.regions=colorRampPalette(rev(brewer.pal(5,"Spectral"))),
+  #     scales=list(draw=FALSE ))
+
+```
+
+```{r rastermap_observations_seasonal_exp, message=FALSE, echo=FALSE}
+#   
+# 
+#   
+#   
+  # ao_data_rasterized_winter <- ratify(ao_data_rasterized_winter)
+  # rat <- levels(ao_data_rasterized_winter)[[1]]
+  # rat$season <- "Winter"
+  # levels(ao_data_rasterized_winter) <- rat
+  # 
+  # ao_data_rasterized_spring <- ratify(ao_data_rasterized_spring)
+  # rat <- levels(ao_data_rasterized_spring)[[1]]
+  # rat$season <- "Spring"
+  # levels(ao_data_rasterized_spring) <- rat
+  # 
+  # ao_data_rasterized_summer <- ratify(ao_data_rasterized_summer)
+  # rat <- levels(ao_data_rasterized_summer)[[1]]
+  # rat$season <- "Summer"
+  # levels(ao_data_rasterized_summer) <- rat
+  # 
+  # ao_data_rasterized_fall <- ratify(ao_data_rasterized_fall)
+  # rat <- levels(ao_data_rasterized_fall)[[1]]
+  # rat$season <- "Fall"
+  # levels(ao_data_rasterized_fall) <- rat
+  # 
+  # 
+  # ao_season_stack <- stack(ao_data_rasterized_winter, ao_data_rasterized_spring, ao_data_rasterized_summer, ao_data_rasterized_fall)
+
+#   
+#   
+#   
+   # View(as.data.frame(ao_data_rasterized_winter))
+#   
+# 
+  # gplot(ao_data_rasterized_winter) +
+  #   geom_tile(aes(fill = log(value))) +
+  #   facet_wrap(~ name) +
+  #   scale_fill_gradient(low = 'white', high = 'blue') +
+  #   coord_equal()
+# 
+# 
+#   ao_data_rasterized_hm<- raster::rasterize(spTransform(ao_data, CRS("+proj=laea +lat_0=52 +lon_0=10 +x_0=4321000 +y_0=3210000 +ellps=GRS80 +units=m +no_defs")), grid, fun='count', field='season', na.rm=TRUE)
+#     
+#     
+#     
+#     
+# 
+  # country_map <- spTransform(country_map, CRS("+proj=laea +lat_0=52 +lon_0=10 +x_0=4321000 +y_0=3210000 +ellps=GRS80 +units=m +no_defs"))
+#   
+# gplot(ao_season_stack) +
+#     coord_fixed() +
+#     annotation_map(as.data.frame(country_map), fill="gray40", colour="black") +
+#     geom_tile(aes(fill=log(value)), alpha=1) +
+#     annotation_map(as.data.frame(country_map), fill=NA, colour="black") +
+#     theme(panel.background = element_rect(fill="aliceblue"), legend.position = "right") +
+#     scale_fill_distiller(palette = "Spectral", na.value=NA) +
+#     guides(color = guide_legend(override.aes = list(size=4))) +
+#     facet_wrap(~ layer) +
+#     # coord_map(projection = "gnomonic", orientation=c(60, 16, -5)) +
+#     xlab("Longitude") +
+#     ylab("Latitude")
+
+```
+
+## Human population
+
 ```{r map_human_population, messages=FALSE, echo=FALSE}
@@
     coord_fixed() +
-    annotation_map(as.data.frame(country_map), fill="gray40", colour="black") +
+    annotation_map(as.data.frame(country_map), fill="gray50", colour="black") +
     geom_tile(aes(fill=log(value)), alpha=1) +
@@
     theme(panel.background = element_rect(fill="aliceblue"), legend.position = "right") +
-    scale_fill_distiller(palette = "Spectral", na.value=NA) +
+    scale_fill_distiller(palette = "Spectral", na.value=NA, direction=-1) +
     guides(color = guide_legend(override.aes = list(size=4))) +
@@
     ylab("Latitude") +
-    labs(fill = "Log population") +
+    labs(fill = "Inhabitants (log)") +
     ggtitle("Human population density")
@@
 
-Log human population per square km in norway
+Log human population per 10 x 10 km cell in norway
 
@@
 
-The log human population per `r gridsizeinkm` x `r gridsizeinkm` km cs cell in Norway.
+The log human population in Norway, rasterized to `r gridsizeinkm` x `r gridsizeinkm` km cells.
 
@@
 
-The number of `r taxon['canonicalName']` in Norway per `r gridsizeinkm` x `r gridsizeinkm` km cs cells, versus the human population in those cells.
-R² of the linaer regression is `r summary(lm(observations ~ population, data=num_vs_pop))$r.squared`.
+The number of observations in Norway within the `r tolower(taxon['rank'])` `r taxon['canonicalName']` per `r gridsizeinkm` x `r gridsizeinkm` km cells, versus the human population in those cells.
+R² of the linear regression is `r summary(lm(observations ~ population, data=num_vs_pop))$r.squared`.
 
@@
     ylab("Latitude") +
-    labs(fill = "Log observations") +
+    labs(fill = "Observations (log)") +
     ggtitle("Observations per capita")
@@
 
-The log number of `r taxon['canonicalName']` observations in Norway per capita within `r gridsizeinkm` x `r gridsizeinkm` km cells.
+The log number of observations in Norway within the `r tolower(taxon['rank'])` `r taxon['canonicalName']`, rasterized to `r gridsizeinkm` x `r gridsizeinkm` km cells.
 
@@
     ylab("Latitude") +
-    labs(fill = "Log observations") +
-    ggtitle("Observations relative to regression with human population")
+    labs(fill = "Observations (log)") +
+    ggtitle("Observations corrected for inhabitants")
 ```
 
-The log number of `r taxon['canonicalName']` observations in Norway divided by the number predicted by the regression line (numberOfObservations = `r B0` + `r B1` * humanPopulation),  within `r gridsizeinkm` x `r gridsizeinkm` km cells.
+The log number of observations within the `r tolower(taxon['rank'])` `r taxon['canonicalName']` in Norway, divided by the number predicted by the regression line (numberOfObservations = `r B0` + `r B1` * humanPopulation), rasterized to `r gridsizeinkm` x `r gridsizeinkm` km cells.
+
+## Bioclimatic variables
+
+```{r climate_data, messages=FALSE, echo=FALSE}
+
+env <- getData('worldclim', var='bio', res=5)
+
+env2 <- projectRaster(env[[1]], grid, method="bilinear")
+
+# plot(env, 1, main=NULL, axes=FALSE)
+
+gplot(env2) +
+    coord_fixed() +
+    annotation_map(as.data.frame(country_map), fill="gray50", colour="black") +
+    geom_tile(aes(fill=value/10), alpha=1) +
+    annotation_map(as.data.frame(country_map), fill=NA, colour="black") +
+    theme(panel.background = element_rect(fill="aliceblue"), legend.position = "right") +
+    scale_fill_distiller(palette = "Spectral", na.value=NA, direction=-1) +
+    guides(color = guide_legend(override.aes = list(size=4))) +
+    # coord_map(projection = "gnomonic", orientation=c(60, 16, -5)) +
+    xlab("Longitude") +
+    ylab("Latitude") +
+    labs(fill = "°C") +
+    ggtitle("Annual Mean Temperature")
+
+```
+
+The Annual Mean Temperature in Norway, rasterized to `r gridsizeinkm` x `r gridsizeinkm` km cells.

### 2018-04-16 exploration.R

@@ -37,8 +37,8 @@ load_GBIF_data <- function(download_filter) {
   file_age = as.numeric((difftime(as.Date(file.info(destfile)$ctime), Sys.Date(), units = "days")))
   
   
-  # If the GBIF data has been deleted, is older than 10 days or the filter has been changed, download new data
-  if(!file.exists(destfile) || file_age > 10 || paste(download_filter, collapse=', ') != paste(readChar(filter_file, file.info(filter_file)$size))) {
+  # If the GBIF data has been deleted, is older than 10 days or the filter has been changed or deleted, download new data
+  if(!file.exists(destfile) || !file.exists(filter_file) || file_age > 10 || paste(download_filter, collapse=', ') != paste(readChar(filter_file, file.info(filter_file)$size))) {
     cat(paste(gbif_filter, collapse=', '), file=filter_file, sep="")
     
     # Get the credentials if needed
@@ -47,12 +47,17 @@ load_GBIF_data <- function(download_filter) {
     }
     
     download_key <- do.call(occ_download, download_filter) %>% occ_download_meta
-    download_GBIF_API(download_key=download_key,destfile_name=destfile,n_try=20,Sys.sleep_duration=30)
+    download_GBIF_API(download_key=download_key,destfile_name=destfile,n_try=40,Sys.sleep_duration=30)
   }
 
-  gbif_data <- read.table(unzip(destfile,files="occurrence.txt"), header=T, sep="\t", quote="", fill=FALSE)[c('gbifID', 'license','year','month','day','decimalLatitude', 'decimalLongitude','coordinateUncertaintyInMeters', 'kingdom', 'phylum', 'class', 'order', 'family', 'genus', 'specificEpithet', 'species', 'taxonRank')]
-  
+  gbif_data <- read.table(unzip(destfile,files="occurrence.txt"), header=T, sep="\t", quote="", fill=FALSE)[c('gbifID', 'recordedBy', 'license','year','month','day','decimalLatitude', 'decimalLongitude','coordinateUncertaintyInMeters', 'kingdom', 'phylum', 'class', 'order', 'family', 'genus', 'specificEpithet', 'species', 'taxonRank')]
+
   gbif_data$season <- date_to_season(paste0(gbif_data$month, "-", gbif_data$day))
[…]

### 2018-05-09 absences.Rmd
Plot species and observations per pseudoevent

@@
   sp::proj4string(country_map) <- CRS("+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs +towgs84=0,0,0")
+  country_map <- spTransform(country_map, CRS("+proj=laea +lat_0=52 +lon_0=10 +x_0=4321000 +y_0=3210000 +ellps=GRS80 +units=m +no_defs"))
 ```
@@
 ```{r variables, include=FALSE, echo=FALSE}
-  higherLevel <- "Family"
+  higherLevel <- "Class"
   focalSpecies <- "Haliaeetus albicilla"
 
-  gridsizeinkm <- 50
+  gridsizeinkm <- 20
   grid <- get_grid(gridsizeinkm, country_coordinates)
@@
 
-  per_user <- ao_data %>% group_by(recordedBy, species) %>% tally %>% tidyr::spread(species, n, fill = 0)
-  per_user$observations <- (ao_data %>% group_by(recordedBy) %>% tally)$n
-
-  per_user$species <- rowSums(per_user[c(-1)] > 0) - 1
-  per_user <- per_user[c("recordedBy",focalSpecies,"observations","species")]
-
-  names(per_user) <- c("recordedBy","observationsOfSpecies","totalObservations","numberOfSpeciesObserved")
+  # per_user <- ao_data %>% group_by(recordedBy, species) %>% tally %>% tidyr::spread(species, n, fill = 0)
+  # per_user$observations <- (ao_data %>% group_by(recordedBy) %>% tally)$n
+  # 
+  # per_user$species <- rowSums(per_user[c(-1)] > 0) - 1
+  # per_user <- per_user[c("recordedBy",focalSpecies,"observations","species")]
+  # 
+  # names(per_user) <- c("recordedBy","observationsOfSpecies","totalObservations","numberOfSpeciesObserved")
 
@@
 
-```{r get_observers, include=FALSE, echo=FALSE}
-  # Get all people that have reported this species, and at least x observations and y species
-  per_user <- per_user %>% filter(observationsOfSpecies > 0 & totalObservations > 100 & numberOfSpeciesObserved > 3)
+```{r ao_data_to_wide, include=FALSE, echo=FALSE}
+
+ao_data_wide <- ao_data[c("decimalLatitude","decimalLongitude","year","month","day","recordedBy","species")]
+
+ao_data_wide <- spread(count(ao_data_wide, decimalLatitude, decimalLongitude, year, month, day, recordedBy, species), species, n, fill = 0)
 
@@
 
-```{r get_empty_squares, include=FALSE, echo=FALSE}
-  # For this species: get cells where it has been observed
-
-  focus_species_observations <- ao_data %>% filter(species == focalSpecies)
-
-  sp::coordinates(focus_species_observations) <- ~decimalLongitude + decimalLatitude
-  sp::proj4string(focus_species_observations) <- CRS("+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs +towgs84=0,0,0")
-  focus_species_observations <- spTransform(focus_species_observations, CRS("+proj=laea +lat_0=52 +lon_0=10 +x_0=4321000 +y_0=3210000 +ellps=GRS80 +units=m +no_defs"))
-   focus_species_cells <- as.data.frame(focus_species_observations)
-  focus_species_cells$decimalLatitude <- focus_species_cells$decimalLatitude %/% (gridsizeinkm * 1000) * (gridsizeinkm * 1000) + (gridsizeinkm * 500)
-  focus_species_cells$decimalLongitude <- focus_species_cells$decimalLongitude %/% (gridsizeinkm * 1000) * (gridsizeinkm * 1000) + (gridsizeinkm * 500)
-
-  focus_species_cells$cell <- paste(focus_species_cells$decimalLatitude, focus_species_cells$decimalLongitude, sep = "_")
-
-  focus_species_cells <- focus_species_cells[c("cell")]
-  focus_species_cells <- unique(focus_species_cells)
+```{r get_sums, message=FALSE, echo=FALSE}
+counted_species <- rowSums(ao_data_wide[,(7:ncol(ao_data_wide))] > 0)
+counted_observations <- rowSums(ao_data_wide[,(7:ncol(ao_data_wide))])
 
@@
 
-```{r get_absence, include=FALSE, echo=FALSE}
-  # For such cells, find the number of filtered people that have been there
-  # Or the number of species reported there by such filtered people
+```{r plot_species_counts, message=FALSE, echo=FALSE}
+ao_data_wide$species <- counted_species
+species_counts <- table(ao_data_wide$species)
+barplot(species_counts, main="Number of species per pseudoevent", xlab="Number of species")
+```
 
-  other_species_observations <- ao_data %>% filter(species != focalSpecies)
-  other_species_observations$date <- paste(other_species_observations$year, other_species_observations$month, other_species_observations$day, sep = "-")
+```{r plot_observation_counts, message=FALSE, echo=FALSE}
+ao_data_wide$observations <- counted_observations
+observation_counts <- table(ao_data_wide$observations)
+barplot(observation_counts, main="Number of observations per pseudoevent", xlab="Number of observations")
+```
 
-  sp::coordinates(other_species_observations) <- ~decimalLongitude + decimalLatitude
-  sp::proj4string(other_species_observations) <- CRS("+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs +towgs84=0,0,0")
-  other_species_observations <- spTransform(other_species_observations, CRS("+proj=laea +lat_0=52 +lon_0=10 +x_0=4321000 +y_0=3210000 +ellps=GRS80 +units=m +no_defs"))
+```{r snap_observations_to_raster, include=FALSE, echo=FALSE}
 
-  other_species_observations <- as.data.frame(other_species_observations)
-  other_species_observations$decimalLatitude <- other_species_observations$decimalLatitude %/% (gridsizeinkm * 1000) * (gridsizeinkm * 1000) + (gridsizeinkm * 500)
-  other_species_observations$decimalLongitude <- other_species_observations$decimalLongitude %/% (gridsizeinkm * 1000) * (gridsizeinkm * 1000) + (gridsizeinkm * 500)
+ao_data_snapped_to_grid <- ao_data
 
-  other_species_observations <- other_species_observations[c("recordedBy","decimalLatitude","decimalLongitude","date")]
-  other_species_observations <- unique(other_species_observations)
-
-  other_species_observations$cell <- paste(other_species_observations$decimalLatitude, other_species_observations$decimalLongitude, sep = "_")
-
-  other_species_observations <- subset(other_species_observations, recordedBy %in% per_user$recordedBy)
-  other_species_observations <- subset(other_species_observations, !(cell %in% focus_species_cells$cell))
-
-  sp::coordinates(other_species_observations) <- ~decimalLongitude + decimalLatitude
-  sp::proj4string(other_species_observations) <- CRS("+proj=laea +lat_0=52 +lon_0=10 +x_0=4321000 +y_0=3210000 +ellps=GRS80 +units=m +no_defs")
-
-  other_species_rasterized <- raster::rasterize(spTransform(other_species_observations, CRS("+proj=laea +lat_0=52 +lon_0=10 +x_0=4321000 +y_0=3210000 +ellps=GRS80 +units=m +no_defs")), grid, fun='count')
-
-  country_map <- spTransform(country_map, CRS("+proj=laea +lat_0=52 +lon_0=10 +x_0=4321000 +y_0=3210000 +ellps=GRS80 +units=m +no_defs"))
+sp::coordinates(ao_data_snapped_to_grid) <- ~decimalLongitude + decimalLatitude
+sp::proj4string(ao_data_snapped_to_grid) <- CRS("+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs +towgs84=0,0,0")
+ao_data_snapped_to_grid <- spTransform(ao_data_snapped_to_grid, CRS("+proj=laea +lat_0=52 +lon_0=10 +x_0=4321000 +y_0=3210000 +ellps=GRS80 +units=m +no_defs"))
+ao_data_snapped_to_grid <- as.data.frame(ao_data_snapped_to_grid)
+ao_data_snapped_to_grid$decimalLatitude <- ao_data_snapped_to_grid$decimalLatitude %/% (gridsizeinkm * 1000) * (gridsizeinkm * 1000) + (gridsizeinkm * 500)
+ao_data_snapped_to_grid$decimalLongitude <- ao_data_snapped_to_grid$decimalLongitude %/% (gridsizeinkm * 1000) * (gridsizeinkm * 1000) + (gridsizeinkm * 500)
 
@@
 
-```{r plots_presence, message=FALSE, echo=FALSE}
+```{r snap_observations_to_raster, include=FALSE, echo=FALSE}
 
-  focus_species_rasterized <- raster::rasterize(spTransform(focus_species_observations, CRS("+proj=laea +lat_0=52 +lon_0=10 +x_0=4321000 +y_0=3210000 +ellps=GRS80 +units=m +no_defs")), grid, fun='count')
+ao_data_snapped_to_grid_wide <- ao_data_snapped_to_grid[c("decimalLatitude","decimalLongitude","year","month","day","recordedBy","species")]
 
-  gplot(focus_species_rasterized) +
-    coord_fixed() +
-    annotation_map(as.data.frame(country_map), fill="grey50", colour="black") +
-    geom_tile(aes(fill=value), alpha=1) +
-    annotation_map(as.data.frame(country_map), fill=NA, colour="black") +
-    theme(panel.background = element_rect(fill="aliceblue"), legend.position = "right") +
-    # scale_fill_continuous(high = "red", low = "white", na.value=NA) +
-    guides(color = guide_legend(override.aes = list(size=4))) +
-    # coord_map(projection = "gnomonic", orientation=c(60, 16, -5)) +
-    xlab("Longitude") +
-    ylab("Latitude") +
-    labs(fill = "Observations") + 
-    ggtitle("Presence")
-```
-
-Observations of _`r focalSpecies`_ in Norway.
-
-```{r plots_absence, message=FALSE, echo=FALSE}
-
-  gplot(other_species_rasterized) +
-    coord_fixed() +
-    annotation_map(as.data.frame(country_map), fill="grey50", colour="black") +
-    geom_tile(aes(fill=value), alpha=1) +
-    annotation_map(as.data.frame(country_map), fill=NA, colour="black") +
-    theme(panel.background = element_rect(fill="aliceblue"), legend.position = "right") +
-    # scale_fill_continuous(high = "red", low = "white", na.value=NA) +
-    guides(color = guide_legend(override.aes = list(size=4))) +
-    # coord_map(projection = "gnomonic", orientation=c(60, 16, -5)) +
-    xlab("Longitude") +
-    ylab("Latitude") +
-    labs(fill = "Absences") + 
-    ggtitle("Absence")
+ao_data_snapped_to_grid_wide <- spread(count(ao_data_snapped_to_grid_wide, decimalLatitude, decimalLongitude, year, month, day, recordedBy, species), species, n, fill = 0)
 
@@
 
-Absence of _`r focalSpecies`_ in Norway.
-(Visits by users known to report _`r focalSpecies`_, where that user reported at least one observation of other species within the same `r tolower(higherLevel)` within that cell.)
+```{r get_sums_snapped, message=FALSE, echo=FALSE}
+counted_species_snapped <- rowSums(ao_data_snapped_to_grid_wide[,(7:ncol(ao_data_snapped_to_grid_wide))] > 0)
+
+counted_observations_snapped <- rowSums(ao_data_snapped_to_grid_wide[,(7:ncol(ao_data_snapped_to_grid_wide))])
+
+```
+
+```{r plot_snapped_species_counts, message=FALSE, echo=FALSE}
+ao_data_snapped_to_grid_wide$species <- counted_species_snapped
+species_counts_snapped <- table(ao_data_snapped_to_grid_wide$species)
+barplot(species_counts_snapped, main="Number of species per pseudoevent, snapped to grid", xlab="Number of species")
+```
+
+```{r plot_snapped_observation_counts, message=FALSE, echo=FALSE}
+ao_data_snapped_to_grid_wide$observations <-counted_observations_snapped
+observation_counts_snapped <- table(ao_data_snapped_to_grid_wide$observations)
+barplot(observation_counts_snapped, main="Number of observations per pseudoevent, snapped to grid", xlab="Number of observations")
+```
+
+<!-- ```{r get_observers, include=FALSE, echo=FALSE} -->
+<!--   # Get all people that have reported this species, and at least x observations and y species -->
+<!--   per_user <- per_user %>% filter(observationsOfSpecies > 0 & totalObservations > 100 & numberOfSpeciesObserved > 3) -->
+
+<!-- ``` -->
+
+<!-- ```{r get_presence_observations , include=FALSE, echo=FALSE} -->
+<!--   # Get one observation per observer per day per rastercell of the focal species -->
+<!--   presence_observations <- ao_data %>% filter(species == focalSpecies) -->
+<!--   presence_observations <- subset(presence_observations, recordedBy %in% per_user$recordedBy) -->
+
+<!--   sp::coordinates(presence_observations) <- ~decimalLongitude + decimalLatitude -->
+<!--   sp::proj4string(presence_observations) <- CRS("+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs +towgs84=0,0,0") -->
+<!--   presence_observations <- spTransform(presence_observations, CRS("+proj=laea +lat_0=52 +lon_0=10 +x_0=4321000 +y_0=3210000 +ellps=GRS80 +units=m +no_defs")) -->
+<!--   presences <- as.data.frame(presence_observations) -->
+<!--   presences$decimalLatitude <- presences$decimalLatitude %/% (gridsizeinkm * 1000) * (gridsizeinkm * 1000) + (gridsizeinkm * 500) -->
+<!--   presences$decimalLongitude <- presences$decimalLongitude %/% (gridsizeinkm * 1000) * (gridsizeinkm * 1000) + (gridsizeinkm * 500) -->
+
+<!--   presences$cell <- paste(presences$decimalLatitude, presences$decimalLongitude, sep = "_") -->
+
+<!--   presences <- presences[c("decimalLatitude","decimalLongitude","year","month","day","recordedBy")] -->
+<!--   presences <- unique(presences) -->
+<!--   presences$summary <- paste( -->
+<!--                                             presences$decimalLatitude, -->
+<!--                                             presences$decimalLongitude, -->
+<!--                                             presences$year, -->
+<!--                                             presences$month, -->
+<!--                                             presences$day, -->
+<!--                                             presences$recordedBy, -->
+<!--                                             sep = "_") -->
+
+<!-- ``` -->
+
+<!-- ```{r get_other_observations , include=FALSE, echo=FALSE} -->
+<!--   # Get one observation per observer per day per rastercell of the non-focal species -->
+<!--   other_observations <- ao_data %>% filter(species != focalSpecies) -->
+<!--   other_observations <- subset(other_observations, recordedBy %in% per_user$recordedBy) -->
+
+<!--   sp::coordinates(other_observations) <- ~decimalLongitude + decimalLatitude -->
+<!--   sp::proj4string(other_observations) <- CRS("+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs +towgs84=0,0,0") -->
+<!--   other_observations <- spTransform(other_observations, CRS("+proj=laea +lat_0=52 +lon_0=10 +x_0=4321000 +y_0=3210000 +ellps=GRS80 +units=m +no_defs")) -->
+<!--   other_observations_cells <- as.data.frame(other_observations) -->
+<!--   other_observations_cells$decimalLatitude <- other_observations_cells$decimalLatitude %/% (gridsizeinkm * 1000) * (gridsizeinkm * 1000) + (gridsizeinkm * 500) -->
+<!--   other_observations_cells$decimalLongitude <- other_observations_cells$decimalLongitude %/% (gridsizeinkm * 1000) * (gridsizeinkm * 1000) + (gridsizeinkm * 500) -->
+
+<!--   other_observations_cells$cell <- paste(other_observations_cells$decimalLatitude, other_observations_cells$decimalLongitude, sep = "_") -->
+
+<!--   other_observations_cells <- other_observations_cells[c("decimalLatitude","decimalLongitude","year","month","day","recordedBy")] -->
+<!--   other_observations_cells <- unique(other_observations_cells) -->
+<!--   other_observations_cells$summary <- paste( -->
+<!--                                             other_observations_cells$decimalLatitude, -->
+<!--                                             other_observations_cells$decimalLongitude, -->
+<!--                                             other_observations_cells$year, -->
+<!--                                             other_observations_cells$month, -->
+<!--                                             other_observations_cells$day, -->
+<!--                                             other_observations_cells$recordedBy, -->
+<!--                                             sep = "_") -->
+
+<!-- ``` -->
+
+<!-- ```{r get_absences , include=FALSE, echo=FALSE} -->
+<!--   # Get absences -->
+<!--   absences <- subset(other_observations_cells, !(summary %in% presences$summary)) -->
+
+<!--   sp::coordinates(absences) <- ~decimalLongitude + decimalLatitude -->
+<!--   sp::proj4string(absences) <- CRS("+proj=laea +lat_0=52 +lon_0=10 +x_0=4321000 +y_0=3210000 +ellps=GRS80 +units=m +no_defs") -->
+
+<!--   sp::coordinates(presences) <- ~decimalLongitude + decimalLatitude -->
+<!--   sp::proj4string(presences) <- CRS("+proj=laea +lat_0=52 +lon_0=10 +x_0=4321000 +y_0=3210000 +ellps=GRS80 +units=m +no_defs") -->
+
+<!-- ``` -->
+
+<!-- ```{r plots_absence, message=FALSE, echo=FALSE} -->
+
+<!--   absences_rasterized <- raster::rasterize(spTransform(absences, CRS("+proj=laea +lat_0=52 +lon_0=10 +x_0=4321000 +y_0=3210000 +ellps=GRS80 +units=m +no_defs")), grid, fun='count') -->
+
+<!--   gplot(absences_rasterized) + -->
+<!--     coord_fixed() + -->
+<!--     annotation_map(as.data.frame(country_map), fill="white", colour="grey50") + -->
+<!--     geom_tile(aes(fill=log(value)), alpha=1) + -->
+<!--     scale_fill_gradient(low = "yellow", high = "red", na.value = NA) + -->
+<!--     theme(panel.background = element_rect(fill="aliceblue"), legend.position = "right") + -->
+<!--     guides(color = guide_legend(override.aes = list(size=4))) + -->
+<!-- #    annotation_map(as.data.frame(country_map), fill=NA, colour="black") + -->
+<!--     xlab("Longitude") + -->
+<!--     ylab("Latitude") + -->
+<!--     labs(fill = "Absences") + -->
+<!--     ggtitle("Absence") -->
+
+<!-- ``` -->
+
+<!-- ```{r plots_presence, message=FALSE, echo=FALSE} -->
+<!--   presences_rasterized <- raster::rasterize(spTransform(presences, CRS("+proj=laea +lat_0=52 +lon_0=10 +x_0=4321000 +y_0=3210000 +ellps=GRS80 +units=m +no_defs")), grid, fun='count') -->
+
+<!--   gplot(presences_rasterized) + -->
+<!--     coord_fixed() + -->
+<!--     annotation_map(as.data.frame(country_map), fill="white", colour="grey50") + -->
+<!--     geom_tile(aes(fill=log(value)), alpha=1) + -->
+<!--     scale_fill_gradient(low = "yellow", high = "darkblue", na.value = NA) + -->
+<!-- #    annotation_map(as.data.frame(country_map), fill=NA, colour="black") + -->
+<!--     theme(panel.background = element_rect(fill="aliceblue"), legend.position = "right") + -->
+<!--     guides(color = guide_legend(override.aes = list(size=4))) + -->
+<!--     xlab("Longitude") + -->
+<!--     ylab("Latitude") + -->
+<!--     labs(fill = "Presences") + -->
+<!--     ggtitle("Presence") -->
+
+<!-- ``` -->
+
+<!-- ```{r plots_presence, message=FALSE, echo=FALSE} -->
+
+<!--   presence_minus_absence <- overlay(presences_rasterized, absences_rasterized, fun=function(r1, r2){return(r1/r2)}) -->
+
+<!--   gplot(presence_minus_absence) + -->
+<!--     coord_fixed() + -->
+<!--     annotation_map(as.data.frame(country_map), fill="white", colour="grey50") + -->
+<!--     geom_tile(aes(fill=log(value)), alpha=1) + -->
+<!--     scale_fill_gradient2(low = "red", high = "darkblue", mid = "yellow", na.value = NA) + -->
+<!--     theme(panel.background = element_rect(fill="aliceblue"), legend.position = "right") + -->
+<!--     guides(color = guide_legend(override.aes = list(size=4))) + -->
+<!--     xlab("Longitude") + -->
+<!--     ylab("Latitude") + -->
+<!--     labs(fill = "log(Presences / absences)") + -->
+<!--     ggtitle("Relative absence") -->
+
+<!-- ``` -->
+
+<!-- ```{r get_empty_squares, include=FALSE, echo=FALSE} -->
+<!--   # For this species: get cells where it has been observed -->
+
+<!--   # focus_species_observations <- ao_data %>% filter(species == focalSpecies) -->
+<!--   #  -->
+<!--   # sp::coordinates(focus_species_observations) <- ~decimalLongitude + decimalLatitude -->
+<!--   # sp::proj4string(focus_species_observations) <- CRS("+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs +towgs84=0,0,0") -->
+<!--   # focus_species_observations <- spTransform(focus_species_observations, CRS("+proj=laea +lat_0=52 +lon_0=10 +x_0=4321000 +y_0=3210000 +ellps=GRS80 +units=m +no_defs")) -->
+<!--   #  focus_species_cells <- as.data.frame(focus_species_observations) -->
+<!--   # focus_species_cells$decimalLatitude <- focus_species_cells$decimalLatitude %/% (gridsizeinkm * 1000) * (gridsizeinkm * 1000) + (gridsizeinkm * 500) -->
+<!--   # focus_species_cells$decimalLongitude <- focus_species_cells$decimalLongitude %/% (gridsizeinkm * 1000) * (gridsizeinkm * 1000) + (gridsizeinkm * 500) -->
+<!--   #  -->
+<!--   # focus_species_cells$cell <- paste(focus_species_cells$decimalLatitude, focus_species_cells$decimalLongitude, sep = "_") -->
+<!--   #  -->
+<!--   # focus_species_cells <- focus_species_cells[c("cell")] -->
+<!--   # focus_species_cells <- unique(focus_species_cells) -->
+
+<!-- ``` -->
+
+<!-- ```{r get_absence, include=FALSE, echo=FALSE} -->
+<!--   # For such cells, find the number of filtered people that have been there -->
+<!--   # Or the number of species reported there by such filtered people -->
+
+<!--   # other_species_observations <- ao_data %>% filter(species != focalSpecies) -->
+<!--   # other_species_observations$date <- paste(other_species_observations$year, other_species_observations$month, other_species_observations$day, sep = "-") -->
+<!--   #  -->
+<!--   # sp::coordinates(other_species_observations) <- ~decimalLongitude + decimalLatitude -->
+<!--   # sp::proj4string(other_species_observations) <- CRS("+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs +towgs84=0,0,0") -->
+<!--   # other_species_observations <- spTransform(other_species_observations, CRS("+proj=laea +lat_0=52 +lon_0=10 +x_0=4321000 +y_0=3210000 +ellps=GRS80 +units=m +no_defs")) -->
+<!--   #  -->
+<!--   # other_species_observations <- as.data.frame(other_species_observations) -->
+<!--   # other_species_observations$decimalLatitude <- other_species_observations$decimalLatitude %/% (gridsizeinkm * 1000) * (gridsizeinkm * 1000) + (gridsizeinkm * 500) -->
+<!--   # other_species_observations$decimalLongitude <- other_species_observations$decimalLongitude %/% (gridsizeinkm * 1000) * (gridsizeinkm * 1000) + (gridsizeinkm * 500) -->
+<!--   #  -->
+<!--   # other_species_observations <- other_species_observations[c("recordedBy","decimalLatitude","decimalLongitude","date")] -->
+<!--   # other_species_observations <- unique(other_species_observations) -->
+<!--   #  -->
+<!--   # other_species_observations$cell <- paste(other_species_observations$decimalLatitude, other_species_observations$decimalLongitude, sep = "_") -->
+<!--   #  -->
+<!--   # other_species_observations <- subset(other_species_observations, recordedBy %in% per_user$recordedBy) -->
+<!--   # other_species_observations <- subset(other_species_observations, !(cell %in% focus_species_cells$cell)) -->
+<!--   #  -->
+<!--   #  -->
+<!--   # sp::coordinates(other_species_observations) <- ~decimalLongitude + decimalLatitude -->
+<!--   # sp::proj4string(other_species_observations) <- CRS("+proj=laea +lat_0=52 +lon_0=10 +x_0=4321000 +y_0=3210000 +ellps=GRS80 +units=m +no_defs") -->
+<!--   #  -->
+<!--   # other_species_rasterized <- raster::rasterize(spTransform(other_species_observations, CRS("+proj=laea +lat_0=52 +lon_0=10 +x_0=4321000 +y_0=3210000 +ellps=GRS80 +units=m +no_defs")), grid, fun='count') -->
+<!--   #  -->
+<!--   #  -->
+<!--   # country_map <- spTransform(country_map, CRS("+proj=laea +lat_0=52 +lon_0=10 +x_0=4321000 +y_0=3210000 +ellps=GRS80 +units=m +no_defs")) -->
+
+<!-- ```   -->
+
+<!-- ```{r plots_presence, message=FALSE, echo=FALSE} -->
+
+<!--   # focus_species_rasterized <- raster::rasterize(spTransform(focus_species_observations, CRS("+proj=laea +lat_0=52 +lon_0=10 +x_0=4321000 +y_0=3210000 +ellps=GRS80 +units=m +no_defs")), grid, fun='count') -->
+<!--   #  -->
+<!--   # gplot(focus_species_rasterized) + -->
+<!--   #   coord_fixed() + -->
+<!--   #   annotation_map(as.data.frame(country_map), fill="grey50", colour="black") + -->
+<!--   #   geom_tile(aes(fill=value), alpha=1) + -->
+<!--   #   annotation_map(as.data.frame(country_map), fill=NA, colour="black") + -->
+<!--   #   theme(panel.background = element_rect(fill="aliceblue"), legend.position = "right") + -->
+<!--   #   # scale_fill_continuous(high = "red", low = "white", na.value=NA) + -->
+<!--   #   guides(color = guide_legend(override.aes = list(size=4))) + -->
+<!--   #   # coord_map(projection = "gnomonic", orientation=c(60, 16, -5)) + -->
+<!--   #   xlab("Longitude") + -->
+<!--   #   ylab("Latitude") + -->
+<!--   #   labs(fill = "Observations") +  -->
+<!--   #   ggtitle("Presence") -->
+<!-- ``` -->
+
+<!-- Observations of _`r focalSpecies`_ in Norway. -->
+
+<!-- ```{r plots_absence, message=FALSE, echo=FALSE} -->
+
+<!--   # gplot(other_species_rasterized) + -->
+<!--   #   coord_fixed() + -->
+<!--   #   annotation_map(as.data.frame(country_map), fill="grey50", colour="black") + -->
+<!--   #   geom_tile(aes(fill=value), alpha=1) + -->
+<!--   #   annotation_map(as.data.frame(country_map), fill=NA, colour="black") + -->
+<!--   #   theme(panel.background = element_rect(fill="aliceblue"), legend.position = "right") + -->
+<!--   #   # scale_fill_continuous(high = "red", low = "white", na.value=NA) + -->
+<!--   #   guides(color = guide_legend(override.aes = list(size=4))) + -->
+<!--   #   # coord_map(projection = "gnomonic", orientation=c(60, 16, -5)) + -->
+<!--   #   xlab("Longitude") + -->
+<!--   #   ylab("Latitude") + -->
+<!--   #   labs(fill = "Absences") +  -->
+<!--   #   ggtitle("Absence") -->
+
+<!-- ``` -->
+
+<!-- Absence of _`r focalSpecies`_ in Norway. -->
+<!-- (Visits by users known to report _`r focalSpecies`_, where that user reported at least one observation of other species within the same `r tolower(higherLevel)` within that cell.) -->
